---
title: "PTM-SEA: Kinase Activity Inference from Phosphoproteomics Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PTM-SEA: Kinase Activity Inference from Phosphoproteomics Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = FALSE
)
```

## Introduction

PTM-SEA (Post-Translational Modification Signature Enrichment Analysis) uses curated phosphosite signatures from PTMsigDB to infer kinase activities from phosphoproteomics data. This vignette demonstrates three approaches:

1. **GSEA without direction** - ignores ;u/;d suffixes in PTMsigDB
2. **GSEA with direction** - splits pathways into _up and _down sets
3. **ORA with direction** - over-representation analysis on significant sites

We analyze both DPA (Differential PTM Abundance) and DPU (Differential PTM Usage) results.

## Load Data and PTMsigDB

```{r load_data}
library(prophosqua)
library(fgsea)
library(clusterProfiler)
library(enrichplot)

# Load example data
example_path <- here::here("data", "combined_test_diff_example.rds")
example_data <- readRDS(example_path)

cat("Data dimensions:", nrow(example_data), "rows x", ncol(example_data), "columns\n")
cat("Contrasts:", paste(unique(example_data$contrast), collapse = ", "), "\n")
```

```{r download_ptmsigdb}
# Download PTMsigDB signatures
gmt_path <- download_ptmsigdb(species = "mouse", output_dir = tempdir())
pathways_raw <- fgsea::gmtPathways(gmt_path)

# Trim pathways to 11-mer (default) to increase overlap with our data
pathways <- trim_ptmsigdb_pathways(pathways_raw, trim_to = "11")
cat("Loaded", length(pathways), "PTMsigDB signatures (trimmed to 11-mer)\n")
```

## Overlap Statistics

Understanding the overlap between our data and PTMsigDB is crucial for interpreting results.

```{r overlap_stats}
# Our data: unique flanking sequences (trimmed to 11-mer to match pathways)
our_sequences <- unique(toupper(trimws(example_data$SequenceWindow)))
our_sequences_trimmed <- sapply(our_sequences, prophosqua:::trim_flanking_seq, trim_to = 11L)
our_site_ids <- paste0(our_sequences_trimmed, "-p")
n_our_sites <- length(unique(our_site_ids))

# PTMsigDB: unique site IDs (strip ;u/;d for comparison)
ptmsigdb_ids_raw <- unique(unlist(pathways))
ptmsigdb_ids_stripped <- unique(gsub(";[ud]$", "", ptmsigdb_ids_raw))
n_ptmsigdb_sites <- length(ptmsigdb_ids_stripped)

# Overlap
overlap_ids <- intersect(unique(our_site_ids), ptmsigdb_ids_stripped)
n_overlap <- length(overlap_ids)

cat("=== Overlap Statistics (11-mer) ===\n")
cat("Our data:", n_our_sites, "unique flanking sequences\n")
cat("PTMsigDB:", n_ptmsigdb_sites, "unique site IDs\n")
cat("Overlap:", n_overlap, "sites\n")
cat("  -", round(100 * n_overlap / n_our_sites, 2), "% of our sites\n")
cat("  -", round(100 * n_overlap / n_ptmsigdb_sites, 2), "% of PTMsigDB sites\n")
```

The low overlap is expected because:

1. PTMsigDB contains curated kinase-substrate relationships (~1,700 unique sites)
2. Our phosphoproteomics data measures ~20,000+ sites
3. Most measured sites have no known kinase annotation

---

# DPA Analysis (Differential PTM Abundance)

DPA uses raw phosphosite fold changes, which include both true signaling changes and protein abundance effects.

## Prepare DPA Data

```{r prep_dpa}
# Use t-statistic for ranking
prep_dpa <- ptmsea_data_prep(
  data = example_data,
  stat_column = "statistic.site",
  seq_window_col = "SequenceWindow",
  contrast_col = "contrast"
)

cat("Contrasts prepared:\n")
for (name in names(prep_dpa$ranks)) {
  cat("  ", name, ":", length(prep_dpa$ranks[[name]]), "sites\n")
}
```

## DPA: GSEA without Direction

Standard GSEA ignoring the ;u/;d direction suffixes in PTMsigDB.

```{r dpa_gsea_nodirection}
results_dpa_nodir <- run_ptmsea(
  ranks_list = prep_dpa$ranks,
  pathways = pathways,
  min_size = 3,
  max_size = 500,
  n_perm = 1000,
  pvalueCutoff = 0.1
)

cat("Results per contrast:\n")
for (name in names(results_dpa_nodir)) {
  cat("  ", name, ":", nrow(results_dpa_nodir[[name]]@result), "significant (FDR < 0.1)\n")
}
```

```{r dpa_gsea_nodir_plot, fig.width=12, fig.height=8}
merged_dpa_nodir <- merge_result(results_dpa_nodir)
dotplot(merged_dpa_nodir, showCategory = 10,
        title = "DPA GSEA (no direction) - Significant (FDR < 0.1)") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

## DPA: GSEA with Direction

Splits pathways into _up and _down sets, runs ascending and descending.

```{r dpa_gsea_direction}
results_dpa_dir <- run_ptmsea_up_down(
  ranks_list = prep_dpa$ranks,
  pathways = pathways,
  min_size = 3,
  max_size = 500,
  n_perm = 1000,
  pvalueCutoff = 0.1
)

cat("Results:\n")
for (name in names(results_dpa_dir)) {
  cat("  ", name, ":", nrow(results_dpa_dir[[name]]@result), "significant (FDR < 0.1)\n")
}
```

```{r dpa_gsea_dir_plot, fig.width=14, fig.height=10}
merged_dpa_dir <- merge_result(results_dpa_dir)
dotplot(merged_dpa_dir, showCategory = 10,
        title = "DPA GSEA (with direction) - Significant (FDR < 0.1)") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

---

# DPU Analysis (Differential PTM Usage)

DPU uses protein-normalized phosphosite fold changes, representing true stoichiometry changes independent of protein abundance.

## Prepare DPU Data

```{r prep_dpu}
# Use t-statistic for DPU ranking
prep_dpu <- ptmsea_data_prep(
  data = example_data,
  stat_column = "tstatistic_I",
  seq_window_col = "SequenceWindow",
  contrast_col = "contrast"
)

cat("Contrasts prepared:\n")
for (name in names(prep_dpu$ranks)) {
  cat("  ", name, ":", length(prep_dpu$ranks[[name]]), "sites\n")
}
```

## DPU: GSEA without Direction

```{r dpu_gsea_nodirection}
results_dpu_nodir <- run_ptmsea(
  ranks_list = prep_dpu$ranks,
  pathways = pathways,
  min_size = 3,
  max_size = 500,
  n_perm = 1000,
  pvalueCutoff = 0.1
)

cat("Results per contrast:\n")
for (name in names(results_dpu_nodir)) {
  cat("  ", name, ":", nrow(results_dpu_nodir[[name]]@result), "significant (FDR < 0.1)\n")
}
```

```{r dpu_gsea_nodir_plot, fig.width=12, fig.height=8}
merged_dpu_nodir <- merge_result(results_dpu_nodir)
dotplot(merged_dpu_nodir, showCategory = 10,
        title = "DPU GSEA (no direction) - Significant (FDR < 0.1)") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

## DPU: GSEA with Direction

```{r dpu_gsea_direction}
results_dpu_dir <- run_ptmsea_up_down(
  ranks_list = prep_dpu$ranks,
  pathways = pathways,
  min_size = 3,
  max_size = 500,
  n_perm = 1000,
  pvalueCutoff = 0.1
)

cat("Results:\n")
for (name in names(results_dpu_dir)) {
  cat("  ", name, ":", nrow(results_dpu_dir[[name]]@result), "significant (FDR < 0.1)\n")
}
```

```{r dpu_gsea_dir_plot, fig.width=14, fig.height=10}
merged_dpu_dir <- merge_result(results_dpu_dir)
dotplot(merged_dpu_dir, showCategory = 10,
        title = "DPU GSEA (with direction) - Significant (FDR < 0.1)") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

## DPU: ORA with Direction

Over-representation analysis on significant sites for all contrasts. When `fc_col` is provided, sites are split into up-regulated and down-regulated sets for separate ORA tests.

```{r dpu_ora}
# Convert pathways to TERM2GENE format and strip direction suffixes
term2gene <- ptmsigdb_to_term2gene(pathways)
term2gene$gene <- gsub(";[ud]$", "", term2gene$gene)

# Get all PTMsigDB site IDs (stripped, without ;u/;d)
ptmsigdb_sites <- unique(term2gene$gene)

# Run ORA for each contrast (up and down separately)
contrasts <- unique(example_data$contrast)
ora_results <- list()

for (contrast_name in contrasts) {
  contrast_data <- example_data[example_data$contrast == contrast_name, ]

  prep_ora <- ptmsea_ora_prep(
    data = contrast_data,
    ptmsigdb_sites = ptmsigdb_sites,
    score_col = "FDR_I",
    fc_col = "diff_diff",
    threshold = 0.1
  )

  cat(contrast_name, "- Up:", length(prep_ora$enriched_up),
      "Down:", length(prep_ora$enriched_down),
      "Background:", length(prep_ora$background), "\n")

  # ORA for up-regulated sites
  if (length(prep_ora$enriched_up) > 0) {
    ora_up <- enricher(
      gene = prep_ora$enriched_up,
      universe = prep_ora$background,
      TERM2GENE = term2gene,
      pvalueCutoff = 0.1
    )
    if (!is.null(ora_up) && nrow(ora_up@result) > 0) {
      ora_results[[paste0(contrast_name, "_up")]] <- ora_up
    }
  }

  # ORA for down-regulated sites
  if (length(prep_ora$enriched_down) > 0) {
    ora_down <- enricher(
      gene = prep_ora$enriched_down,
      universe = prep_ora$background,
      TERM2GENE = term2gene,
      pvalueCutoff = 0.1
    )
    if (!is.null(ora_down) && nrow(ora_down@result) > 0) {
      ora_results[[paste0(contrast_name, "_down")]] <- ora_down
    }
  }
}
```

```{r dpu_ora_plot, fig.width=14, fig.height=10}
if (length(ora_results) > 0) {
  merged_ora <- merge_result(ora_results)
  if (nrow(merged_ora@compareClusterResult) > 0) {
    print(dotplot(merged_ora, showCategory = 10,
            title = "DPU ORA (up/down separate) - Significant (FDR < 0.1)") +
      ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)))
  } else {
    cat("No significant ORA results at FDR < 0.1\n")
  }
} else {
  cat("No significant ORA results at FDR < 0.1\n")
}
```

---

## Interpretation

- **DPA results** show kinase activity changes including protein abundance effects
- **DPU results** show kinase activity changes independent of protein abundance

Kinases appearing significant in **DPU but not DPA** represent genuine signaling changes.
Kinases appearing in **DPA but not DPU** may be driven by protein abundance changes.

## Session Info

```{r session_info}
sessionInfo()
```
