---
title: "DPA and DPU Integration Analysis"
author:
  - FGCZ
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
params:
  objects_rds: NULL
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
vignette: >
  %\VignetteIndexEntry{DPA and DPU Integration Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  chunk_output_type: console
---

```{r setupglobal, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

suppressPackageStartupMessages({
  library(prophosqua)
  library(dplyr)
})

# The pipeline hands over what its compute step wrote; without it the report
# renders against the example data the package ships.
pipeline_mode <- !is.null(params$objects_rds)
objects <- if (pipeline_mode) {
  readRDS(params$objects_rds)
} else {
  prophosqua:::compute_dpa_dpu_example()
}
```

# Introduction

This analysis integrates total proteome and phosphoproteome DEA results to compute:

- **DPA** (Differential PTM Abundance): Raw phosphosite signal changes
- **DPU** (Differential PTM Usage): Protein-normalized changes (site - protein) reflecting modification stoichiometry

Both were computed by `prophosqua::compute_dpa_dpu()` before this report was
rendered; this document describes and summarizes that result rather than
producing it. Nothing here is re-quantified: the two DEA runs it reads from,

- phospho: `` `r basename(objects$phospho_dea_dir)` ``
- total proteome: `` `r basename(objects$protein_dea_dir)` ``

had already happened.

# DPA: Integrate PTM and Protein Data

Each phosphosite result is paired with the protein-level result of the same
protein and the same contrast, suffixed `.site` and `.protein`. The pairing
gives `r format(objects$n_dpa_rows, big.mark = ",")` site-contrast rows.

## Match Rate

```{r matchrate}
knitr::kable(
  objects$match_rates,
  col.names = c("Contrast", "Total Sites", "Matched Sites", "Match Rate (%)"),
  caption = paste0(
    "How many phosphosites could be paired with a protein-level result, per contrast. Total ",
    "Sites counts the sites tested in the phospho DEA, Matched Sites those whose protein was ",
    "also quantified in the total proteome DEA, and Match Rate their ratio. Only matched ",
    "sites can be corrected for protein abundance, so unmatched sites carry a DPA result but ",
    "no DPU value."
  )
)
```

# DPU: Compute Differential PTM Usage

DPU calculates protein-normalized PTM changes:

- Effect size: $\Delta_I = \log_2FC_{PTM} - \log_2FC_{Protein}$
- Standard error: $SE_I = \sqrt{SE_{PTM}^2 + SE_{Protein}^2}$

The DPU table holds `r format(objects$n_dpu_rows, big.mark = ",")` rows. Only the
rows a site and its protein both contributed to carry a usage estimate; sites
whose protein was not quantified, and proteins for which no site was tested, are
kept alongside them so that nothing is silently dropped, and the `measured_In`
column says which of the three each row is.

# Output Files

| File | Content |
|:-----|:--------|
| `` `r basename(objects$dpa_xlsx)` `` | Differential PTM Abundance, sheet `combinedSiteProteinData` |
| `` `r basename(objects$dpu_xlsx)` `` | Differential PTM Usage, sheet `combinedStats` |
| `combined_test_diff.rds` | the DPU table, as the integration overview reads it |
| `dpa_dpu_objects.rds` | the counts and match rates quoted in this report |

# Session Info

```{r sessioninfo}
sessionInfo()
```
