---
title: "Kinase Activity (Kinase Library + MEA)"
subtitle: "`r params$analysis_type`"
author: "FGCZ"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  objects_rds: NULL
  analysis_type: "DPA"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    code_folding: hide
    self_contained: true
vignette: >
  %\VignetteIndexEntry{Kinase Activity (Kinase Library + MEA)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 8
)

library(dplyr)
library(DT)
library(tidyr)
library(ggplot2)
library(purrr)
library(forcats)
library(writexl)
library(prophosqua)

# Detect pipeline vs vignette mode
pipeline_mode <- !is.null(params$objects_rds)
```

# Overview

**Motif Enrichment Analysis (MEA)** asks, for every kinase, whether the phosphosites that
carry that kinase's preferred sequence motif change *coordinately* in a contrast. It
therefore reports inferred kinase *activity* changes in the **`r params$analysis_type`**
results — not kinase abundance, which a phosphoproteomics experiment does not measure.

## How the numbers in this report are produced

| Step | What happens | Software |
|:--|:--|:--|
| 1. Rank the sites | All quantified sites of a contrast are sorted by their site-level moderated t-statistic (`statistic.site`: most increased first, most decreased last) and written to a `.rnk` file, one per contrast. The sites, the sheet and the ranking statistic come from the combined `PTM_results.xlsx` that PTM-SEA and the KinaseLib GSEA also read, so the three enrichment analyses rank identically. For DPU that statistic is the site-versus-protein interaction, which the combined workbook stores under the same name | `prep_kinaselib.R` of the PTM pipeline |
| 2. Score the motifs | The ±7-residue sequence window of every site is scored against the position-specific scoring matrix of every kinase, and each score is converted into a percentile relative to a reference phosphoproteome | [kinase-library](https://github.com/TheKinaseLibrary/kinase-library) (Python) |
| 3. Build substrate sets | Per kinase, all sites scoring above the configured percentile threshold (95th percentile by default) form its predicted substrate set — one set per kinase (311 Ser/Thr, 78 Tyr kinases in Kinase Library v1.2.0) | kinase-library |
| 4. Test for enrichment | Pre-ranked GSEA of those sets against the ranked site list, giving ES, NES, a permutation p-value (1000 permutations by default) and a Benjamini-Hochberg FDR per kinase | [GSEApy](https://gseapy.readthedocs.io) `prerank`, called by kinase-library |
| 5. This report | Reads the per-contrast `mea_*.csv` files and produces the summaries, figures and Excel export below | R: `prophosqua`, `dplyr`, `ggplot2`, `DT` |

Steps 2-4 run outside R, through the `run-mea` command line entry point of the
kinase-library package. The kinase type, percentile threshold and permutation count
actually used are recorded in the pipeline log written next to the results
(`logs/mea_<contrast>.log`).

## What "enriched at the extremes of the ranked data" means

The ranked list from step 1 is the whole quantified phosphoproteome of one contrast, ordered
by the moderated t-statistic: strongly up-regulated sites at the top, unchanged sites in the
middle, strongly down-regulated sites at the bottom. Ranking on the t-statistic rather than
on the fold change alone means a site needs both a sizeable change and a reproducible one to
reach an extreme of the list. Pre-ranked GSEA walks this list from top to
bottom with a running sum that increases at every site belonging to the kinase's substrate
set and decreases at every site outside it. The largest deviation of that walk from zero is
the **enrichment score (ES)**:

- substrates spread evenly over the list → the walk stays near zero → ES ≈ 0, no signal;
- substrates piling up at the **top** of the list → large **positive** ES: the motif is
  enriched at the up-regulated extreme;
- substrates piling up at the **bottom** → large **negative** ES: enriched at the
  down-regulated extreme.

Randomly permuting the ranks calibrates ES into the **normalized enrichment score (NES)**,
which is comparable between kinases and contrasts, plus a p-value and an FDR. The
**leading-edge substrates** are the sites in front of the running-sum peak: they are the
sites that actually produce the score, and they are listed per kinase in the `mea_*.csv`
files.

So a kinase with NES > 0 and a small FDR has its predicted substrates concentrated among
the sites that go up in this contrast, which is read as increased activity of that kinase;
NES < 0 is read as decreased activity.

Two limitations to keep in mind when interpreting the tables below:

- Substrate sets are **motif predictions**, not experimentally curated kinase-substrate
  pairs. Kinases within a family share near-identical motifs, so their sets overlap heavily
  and their NES values move together — MEA points at a motif class rather than at one
  specific kinase. The PTM-SEA report is the complementary view, based on curated
  PhosphoSitePlus kinase-substrate annotations.
- Only sites with a usable sequence window enter the ranking, and a site's window is scored
  regardless of whether that kinase is expressed in the studied system.

## References and further reading

- The Kinase Library web tool: <https://kinase-library.phosphosite.org>
- kinase-library Python package: <https://github.com/TheKinaseLibrary/kinase-library>
  ([PyPI](https://pypi.org/project/kinase-library/)); this pipeline calls the fork
  <https://github.com/wolski/kinase-library>, which adds the `scan-motifs` and `run-mea`
  command line entry points
- Ser/Thr kinome atlas: Johnson JL *et al.* (2023) An atlas of substrate specificities for
  the human serine/threonine kinome. *Nature* 613:759-766.
  <https://doi.org/10.1038/s41586-022-05575-3>
- Tyrosine kinome atlas: Yaron-Barir TM *et al.* (2024) The intrinsic substrate specificity
  of the human tyrosine kinome. *Nature* 629:1174-1181.
  <https://doi.org/10.1038/s41586-024-07407-y>
- GSEA method: Subramanian A *et al.* (2005) Gene set enrichment analysis. *PNAS*
  102:15545-15550. <https://doi.org/10.1073/pnas.0506580102>
- GSEApy implementation: Fang Z, Liu X, Peltz G (2023) GSEApy: a comprehensive package for
  performing gene set enrichment analysis in Python. *Bioinformatics* 39:btac757.
  <https://doi.org/10.1093/bioinformatics/btac757>

# Load MEA Results

```{r loadMEA}
if (pipeline_mode) {
  mea <- readRDS(params$objects_rds)
} else {
  # Standalone mode: compute from the example data shipped with the package,
  # through the same function the pipeline calls.
  mea <- prophosqua:::compute_mea_example()
}

# The settings the collection step ran with. `params` is a locked binding
# during a render and cannot be extended, so they get their own name.
inputs <- mea$analysis_inputs
output_dir <- inputs$kinaselib_dir

cat("Loaded", nrow(mea$mea_clean), "results from", mea$n_files, "contrasts\n")
cat("Contrasts:", paste(unique(mea$mea_clean$contrast), collapse = ", "), "\n")
```

```{r cleanResults}
mea_clean <- mea$mea_clean
summary_df <- mea$summary_df

knitr::kable(
  summary_df,
  caption = paste0(
    "Number of kinase motif sets tested per contrast and how many of them are ",
    "significantly enriched (", params$analysis_type, " results, GSEA on sites ranked by ",
    "the site-level moderated t-statistic). total_kinases counts the motif sets tested, ",
    "sig_up the kinases with FDR < 0.1 and NES > 0 (predicted substrates enriched among ",
    "up-regulated sites, i.e. inferred activation) and sig_down those with FDR < 0.1 and ",
    "NES < 0 (inferred inactivation)."
  )
)

# Contrast order and per-contrast figure captions, used by the loop below
contrast_ids <- unique(mea_clean$contrast)
dotplot_caps <- paste0(
  "Kinase motif enrichment in contrast ", contrast_ids, " (", params$analysis_type,
  "): normalized enrichment score of the 30 kinases with the smallest FDR. Each point is ",
  "one kinase motif set; the x-axis is the NES from pre-ranked GSEA over sites ranked by ",
  "the site-level moderated t-statistic, so NES > 0 means the predicted substrates ",
  "accumulate among ",
  "up-regulated sites and NES < 0 among down-regulated sites. Point size is -log10(FDR), ",
  "point colour follows the NES (blue negative, red positive), faded points are kinases ",
  "with FDR >= 0.1, and the dashed line marks NES = 0. Kinases are ordered by NES."
)

heatmap_cap <- paste0(
  "Normalized enrichment score per kinase motif set and contrast (", params$analysis_type,
  "). Rows are the 30 kinases with the smallest FDR in any contrast (only kinases reaching ",
  "FDR < 0.15 somewhere are eligible), ordered by NES; columns are the contrasts. Tile ",
  "colour is the NES (blue: substrates enriched among down-regulated sites, red: among ",
  "up-regulated sites, white: no enrichment) and the asterisks give the FDR of that ",
  "kinase in that contrast (* < 0.1, ** < 0.05, *** < 0.01). Empty tiles mean the kinase ",
  "was not tested in that contrast."
)

volcano_cap <- paste0(
  "Enrichment strength versus significance for every tested kinase motif set, one panel ",
  "per contrast (", params$analysis_type, "). Each point is one kinase; the x-axis is the ",
  "NES from pre-ranked GSEA (negative: substrates enriched among down-regulated sites, ",
  "positive: among up-regulated sites) and the y-axis -log10(FDR), with the horizontal ",
  "dashed line at FDR = 0.1 and the vertical dashed line at NES = 0. Colour encodes the ",
  "direction of enrichment, faded points are kinases above the FDR cutoff, and the five ",
  "kinases with the smallest FDR per contrast are labelled. Axes are free per panel."
)
```

# Results by Contrast {.tabset .tabset-pills}

```{r contrastPlots, results='asis', fig.height=8, fig.cap=dotplot_caps}
for (ctr in contrast_ids) {
  cat("\n\n## ", ctr, "\n\n")

  ctr_data <- mea_clean |> filter(contrast == ctr)
  n_sig <- sum(ctr_data$FDR < 0.1, na.rm = TRUE)
  cat("**Significant kinases (FDR < 0.1):** ", n_sig, "\n\n")

  # Top kinases dotplot using shared function
  p <- plot_enrichment_dotplot(
    ctr_data,
    item_col = "kinase",
    fdr_col = "FDR",
    title = paste0(params$analysis_type, " - ", ctr),
    subtitle = "Top 30 kinases by FDR"
  )
  print(p)
  cat("\n\n")

  # Significant kinases table
  cat("### Significant Kinases\n\n")
  sig_table <- ctr_data |>
    filter(FDR < 0.1) |>
    select(kinase, NES, pvalue, FDR, n_leading, set_size) |>
    arrange(FDR) |>
    mutate(across(where(is.numeric), ~round(.x, 4)))
  print(htmltools::tagList(
    DT::datatable(sig_table,
                  extensions = 'Buttons',
                  options = list(pageLength = 15, scrollX = TRUE,
                                 dom = 'Bfrtip', buttons = c('copy', 'csv', 'excel')),
                  caption = paste0(
                    "Kinases enriched at FDR < 0.1 in contrast ", ctr, ", sorted by FDR. ",
                    "NES is the normalized enrichment score (> 0 substrates enriched among ",
                    "up-regulated sites, < 0 among down-regulated sites), pvalue the GSEA ",
                    "permutation p-value, FDR its Benjamini-Hochberg adjustment, ",
                    "set_size the number of ranked sites whose sequence window matches the ",
                    "kinase motif, and n_leading how many of them lie in the leading edge ",
                    "and thus drive the score."
                  ))
  ))
  cat("\n\n")
}
```

# Summary Heatmap

```{r heatmap, fig.height=10, fig.width=12, fig.cap=heatmap_cap}
# Using shared heatmap function
plot_enrichment_heatmap(
  mea_clean,
  item_col = "kinase",
  fdr_col = "FDR",
  n_top = 30,
  title = paste0("MEA Summary Heatmap - ", params$analysis_type)
)
```

# Volcano Plot

```{r volcano, fig.height=8, fig.width=12, fig.cap=volcano_cap}
# Using shared volcano function
plot_enrichment_volcano(
  mea_clean,
  item_col = "kinase",
  fdr_col = "FDR",
  title = paste0("MEA Volcano Plots - ", params$analysis_type)
)
```

# Diagnostics

```{r diagnosticPvalues}
pval_diag <- mea_clean |>
  group_by(contrast) |>
  summarise(
    `Min p-value` = signif(min(pvalue, na.rm = TRUE), 3),
    `p < 0.05` = sum(pvalue < 0.05, na.rm = TRUE),
    `p < 0.01` = sum(pvalue < 0.01, na.rm = TRUE),
    Total = n(),
    .groups = "drop"
  ) |>
  rename(Contrast = contrast)
knitr::kable(
  pval_diag,
  caption = paste0(
    "Distribution of the unadjusted GSEA permutation p-values per contrast (",
    params$analysis_type, "), before FDR adjustment. Total is the number of kinase motif ",
    "sets tested; a contrast with a real kinase signal shows more sets below the ",
    "thresholds than the roughly 5% and 1% of Total expected by chance. The smallest ",
    "attainable p-value is 1/(number of permutations)."
  )
)
```

# All Results

```{r allResults}
# All kinases across all contrasts
all_results_dt <- mea_clean |>
  select(contrast, kinase, NES, pvalue, FDR, n_leading, set_size) |>
  arrange(contrast, FDR) |>
  mutate(across(where(is.numeric), ~round(.x, 4)))

DT::datatable(all_results_dt,
  filter = "top",
  extensions = 'Buttons',
  options = list(pageLength = 15, scrollX = TRUE,
                 dom = 'Bfrtip', buttons = c('copy', 'csv', 'excel')),
  caption = paste0(
    "Complete MEA result table: every tested kinase motif set in every contrast (",
    params$analysis_type, "), sorted by contrast and FDR, including the ",
    "non-significant ones. NES is the normalized enrichment score (> 0 substrates ",
    "enriched among up-regulated sites, < 0 among down-regulated sites), pvalue the ",
    "permutation p-value, FDR its Benjamini-Hochberg adjustment, set_size the number of ",
    "ranked sites matching the kinase motif and n_leading the leading-edge subset that ",
    "drives the score. Use the column filters to search a kinase."
  ))
```

# Export Results

```{r exportedFiles}
stem <- file.path(output_dir, paste0("MEA_", params$analysis_type, "_results"))
exported <- paste0(stem, c(".xlsx", ".rds"))

knitr::kable(
  tibble(
    Output = c("Excel results", "RDS object"),
    File = basename(exported),
    Size = ifelse(
      file.exists(exported),
      paste(round(file.size(exported) / 1024, 1), "KB"),
      "not found"
    )
  ),
  caption = paste0(
    "Files the collection step, prophosqua::compute_mea(), wrote next to this report: the ",
    "Excel workbook with all kinase results, a significant-only sheet and the per-contrast ",
    "summary, and the RDS with the cleaned table this report renders from."
  )
)
```

# Session Info

```{r sessionInfo}
sessionInfo()
```
