---
title: "PTM Analysis - N-to-C Plots"
subtitle: "`r params$analysis_type` Analysis"
author: "FGCZ"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  xlsx_file: NULL
  sheet: "DPA"
  analysis_type: "dpa"
  output_dir: NULL
  fdr: 0.05
  fc: 0.6
  max_fig: 10000
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    code_folding: hide
    self_contained: true
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{N-to-C Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

suppressPackageStartupMessages({
  library(prophosqua)
  library(dplyr)
  library(readxl)
  library(patchwork)
})

# Detect pipeline vs vignette mode
pipeline_mode <- !is.null(params$xlsx_file)

# Load shared utilities if in pipeline mode
if (pipeline_mode && file.exists("src/feature_preparation.R")) {
  source("src/feature_preparation.R")
}

max_fig <- params$max_fig
```

# Introduction

This document generates N-to-C plots for **`r toupper(params$analysis_type)`** analysis.

```{r analysisDescription, results='asis'}
desc <- switch(params$analysis_type,
  "dpa" = "**DPA (Differential PTM Abundance)**: Raw PTM signal changes without protein abundance correction. Shows both site-level and protein-level fold changes.",
  "dpu" = "**DPU (Differential PTM Usage)**: Protein-normalized PTM changes (model-first approach). Reflects genuine modification stoichiometry changes.",
  "cf" = "**CF (CorrectFirst)**: Alternative protein-correction approach where abundances are normalized before statistical modeling."
)
cat(desc)
```

# Data Loading

```{r loadData}
if (pipeline_mode) {
  message("Loading data from: ", params$xlsx_file, " (sheet: ", params$sheet, ")")
  data <- readxl::read_xlsx(params$xlsx_file, sheet = params$sheet)
  output_dir <- if (!is.null(params$output_dir)) params$output_dir else dirname(params$xlsx_file)
} else {
  # Vignette mode: use example data
  data("combined_test_diff_example", package = "prophosqua")
  data <- combined_test_diff_example
  output_dir <- tempdir()
  max_fig <- 5
  message("Using example data from prophosqua package")
}

data_info <- tibble(
  Property = c("Mode", "Sheet", "Analysis Type", "Rows", "Contrasts"),
  Value = c(
    if (pipeline_mode) basename(params$xlsx_file) else "Example data",
    params$sheet, toupper(params$analysis_type),
    nrow(data), paste(unique(data$contrast), collapse = ", ")
  )
)
knitr::kable(
  data_info,
  caption = paste0(
    "Input of this N-to-C report: source workbook, sheet, analysis type, number of ",
    "site-by-contrast rows read and the contrasts they cover. Every row is one phosphosite ",
    "in one contrast, before the significance filter that selects the plotted proteins."
  )
)
```

```{r prepareData}
# Prepare data for plotting using shared function
plot_data <- prepare_ntoc_data(data, params$analysis_type)

all_contrasts <- unique(plot_data$contrast)
cat("Found", length(all_contrasts), "contrasts:", paste(all_contrasts, collapse = ", "), "\n")
```

# Generate N-to-C Plots

**Filtering criteria:** Proteins are included if they contain at least one PTM site with FDR < `r params$fdr` and |log2 fold change| > `r params$fc`.

```{r generatePlots, results='hide'}
if (params$analysis_type == "dpa") {
  # DPA: Use expression plotting (shows protein + site)
  plot_result <- n_to_c_expression_multicontrast(
    plot_data,
    FDR_threshold = params$fdr,
    fc_threshold = params$fc,
    max_plots = max_fig
  )
} else {
  # DPU/CF: Use usage plotting (shows site only)
  plot_result <- n_to_c_usage_multicontrast(
    plot_data,
    FDR_threshold = params$fdr,
    fc_threshold = params$fc,
    max_plots = max_fig
  )
}

n_plots <- nrow(plot_result)
stopifnot("No significant sites found. Adjust fdr/log2fc thresholds in config." = n_plots > 0)
cat("Generated", n_plots, "multi-contrast plots\n")
```

## Example Plots

Showing example plots in the HTML report: one protein WITH protein-level data and one WITHOUT (if available). `r if(pipeline_mode) paste0("All ", n_plots, " plots are exported to PDF.")`

```{r selectExamples}
# Pick the examples first, so that each one can be given its own figure caption.
matched_proteins <- character(0)
protein_note <- NULL

if (params$analysis_type == "dpa") {
  # DPA: show one protein with and one without matched protein-level data
  matched_proteins <- plot_data |>
    dplyr::filter(!is.na(diff.protein)) |>
    dplyr::pull(protein_Id) |>
    unique()

  unmatched_proteins <- plot_data |>
    dplyr::filter(is.na(diff.protein)) |>
    dplyr::pull(protein_Id) |>
    unique()

  matched_idx <- which(plot_result$protein_Id %in% matched_proteins)
  unmatched_idx <- which(plot_result$protein_Id %in% unmatched_proteins)

  example_idx <- c(
    if (length(matched_idx) > 0) matched_idx[1],
    if (length(unmatched_idx) > 0) unmatched_idx[1]
  )
  if (length(example_idx) == 0) {
    example_idx <- seq_len(min(2, n_plots))
  }

  protein_note <- paste0(
    "**Note:** ", length(matched_idx), " proteins have protein-level data, ",
    length(unmatched_idx), " proteins do not."
  )
} else {
  # DPU/CF: no protein-level track to compare against, show the first two
  example_idx <- seq_len(min(2, n_plots))
}

example_has_protein <- plot_result$protein_Id[example_idx] %in% matched_proteins
example_labels <- ifelse(example_has_protein, " (with protein data)", "")
if (params$analysis_type == "dpa") {
  example_labels <- ifelse(example_has_protein, " (with protein data)", " (NO protein data)")
}

example_caps <- paste0(
  "Position-resolved ", toupper(params$analysis_type), " changes along protein ",
  plot_result$protein_Id[example_idx], example_labels,
  ", one panel per contrast. The x-axis is the residue position from the N- to the ",
  "C-terminus of the protein and the y-axis is the site-level log2 fold change of that ",
  "contrast; each vertical stick is one PTM site, coloured by modified residue (S blue, ",
  "T green, Y brown, not localized pink) and dashed when the value was imputed. Asterisks ",
  "give the site FDR (** < 0.05, * < 0.2), and the yellow rectangle spans the protein from ",
  "N to C at the protein-level log2 fold change, so sticks reaching beyond it change more ",
  "than the protein itself. Proteins are shown when at least one site has FDR < ",
  params$fdr, " and |log2FC| > ", params$fc, "."
)
```

`r if (!is.null(protein_note)) protein_note`

```{r displayExamples, fig.width=12, fig.height=10, results='asis', fig.cap=example_caps}
if (n_plots > 0) {
  for (i in example_idx) {
    cat("\n\n### ", plot_result$protein_Id[i], example_labels[match(i, example_idx)], "\n\n", sep = "")
    print(plot_result$plot[[i]])
    cat("\n\n")
  }
} else {
  cat("No significant proteins found with the current filtering criteria.\n")
}
```

# Export Plots

```{r exportPdf, results='hide', eval=pipeline_mode}
# Use explicit output_dir if provided, otherwise use xlsx directory
pdf_dir <- output_dir

if (!dir.exists(pdf_dir)) {
  dir.create(pdf_dir, recursive = TRUE)
}

# Determine output filename based on analysis type
pdf_name <- switch(params$analysis_type,
  "dpa" = "Site_differential_Expression_MultiContrast.pdf",
  "dpu" = "Site_differential_UsageChange_MultiContrast.pdf",
  "cf"  = "Site_differential_CorrectFirst_MultiContrast.pdf"
)

pdf_path <- file.path(pdf_dir, pdf_name)

if (n_plots > 0) {
  pdf(pdf_path, width = 14, height = 10)
  for (i in seq_len(n_plots)) {
    print(plot_result$plot[[i]])
  }
  dev.off()
  cat("Exported", n_plots, "plots to:", pdf_path, "\n")
} else {
  cat("No plots to export.\n")
}
```

```{r exportPdfVignette, results='hide', eval=!pipeline_mode}
message("Vignette mode: PDF export skipped.")
pdf_path <- NULL
```

# Results Summary

```{r summary}
summary_info <- tibble(
  Metric = c("Analysis Type", "Total Proteins", "Plots Generated",
             "Shown in HTML", "FDR Threshold", "FC Threshold"),
  Value = c(toupper(params$analysis_type), n_distinct(plot_data$protein_Id),
            n_plots, min(2, n_plots), params$fdr, params$fc)
)
knitr::kable(
  summary_info,
  caption = paste0(
    "Coverage of this report: how many proteins carry data, how many passed the significance ",
    "filter and were plotted, and how many of those plots are embedded here. Total Proteins ",
    "counts all proteins in the sheet, Plots Generated the proteins with at least one site ",
    "below the FDR threshold and above the fold-change threshold (capped by the max_fig ",
    "setting), and the remaining plots are available in the exported PDF."
  )
)

if (pipeline_mode && n_plots > 0) {
  cat("\nPDF exported to:", pdf_path, "\n")
}
```

# Session Info

```{r sessionInfo}
sessionInfo()
```
