---
title: "PTM Analysis - Sequence Logo"
subtitle: "`r params$sheet` Analysis"
author: "FGCZ"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  xlsx_file: NULL
  sheet: "DPA"
  fdr: 0.05
  fc: 0.6
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    code_folding: hide
    self_contained: true
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Sequence Logo Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

suppressPackageStartupMessages({
  library(dplyr)
  library(ggseqlogo)
  library(readxl)
  library(prophosqua)
})

# Detect pipeline vs vignette mode
pipeline_mode <- !is.null(params$xlsx_file)

# Load shared utilities if in pipeline mode
if (pipeline_mode && file.exists("src/feature_preparation.R")) {
  source("src/feature_preparation.R")
}

is_docx <- knitr::pandoc_to() == "docx"
if (length(is_docx) == 0) {
  is_docx <- TRUE
}
```

# Introduction

This document performs sequence logo analysis on significantly regulated sites from **`r params$sheet`** analysis. Sequence motifs help identify active kinases driving PTM changes.

```{r analysisDescription, results='asis'}
desc <- switch(params$sheet,
  "DPA" = "**DPA (Differential PTM Abundance)**: Raw PTM signal changes. Sequence motifs may reflect both abundance and stoichiometry effects.",
  "DPU" = "**DPU (Differential PTM Usage)**: Protein-normalized changes. Sequence motifs reflect genuine kinase activity changes.",
  "CF" = "**CF (CorrectFirst)**: Alternative normalization. Sequence motifs reflect activity changes with different correction approach."
)
cat(desc)
```

# Data Loading

```{r loadData}
if (pipeline_mode) {
  message("Loading data from: ", params$xlsx_file, " (sheet: ", params$sheet, ")")
  data <- readxl::read_xlsx(params$xlsx_file, sheet = params$sheet)
} else {
  # Vignette mode: use example data
  data("combined_test_diff_example", package = "prophosqua")
  data <- combined_test_diff_example
  message("Using example data from prophosqua package")
}

data_info <- tibble(
  Property = c("Mode", "Sheet", "Rows", "Contrasts"),
  Value = c(
    if (pipeline_mode) basename(params$xlsx_file) else "Example data",
    params$sheet,
    nrow(data), paste(unique(data$contrast), collapse = ", ")
  )
)
knitr::kable(data_info, caption = "Data Summary")
```

# Filter Significant Sites

Filter sites with |log2FC| > `r params$fc` and FDR < `r params$fdr`.

```{r filterSignificant}
# Use shared filtering function with sequence validation
significant_sites <- data |>
  dplyr::filter(!is.na(posInProtein)) |>
  filter_significant_sites(
    fdr_threshold = params$fdr,
    fc_threshold = params$fc,
    require_sequence = TRUE
  )

cat("Found", nrow(significant_sites), "significant sites\n")
```

## Count Significant Sites

```{r tableSignificant}
stopifnot("No significant sites found. Adjust fdr/log2fc thresholds in config." = nrow(significant_sites) > 0)

tx <- as.data.frame(with(significant_sites, table(contrast, regulation, modAA)))
capt <- paste0("Number of significantly regulated sites by contrast, residue, and regulation direction (", params$sheet, ").")

if (is_docx) {
  ft <- flextable::flextable(tx) |>
    flextable::fontsize(part = "all", size = 9) |>
    flextable::set_table_properties(width = 1, layout = "fixed", opts_word = list(split = FALSE)) |>
    flextable::width(width = c(2.5, 1.2, 0.8, 1.0)) |>
    flextable::set_caption(capt)
  ft
} else {
  tx |>
    knitr::kable(caption = capt) |>
    kableExtra::kable_styling(font_size = 9, latex_options = "scale_down")
}
```

## Validate Sequence Window

Ensure the central residue matches the modified amino acid.

```{r validateSeqWindow}
# Use shared validation function
significant_sites <- validate_sequence_window(significant_sites)

cat("After validation:", nrow(significant_sites), "sites remain\n")
```

# Sequence Logo Analysis

## Generate Sequence Logos

```{r figSeqlogo, fig.width=9, fig.height=12, fig.cap="Sequence logos for significantly regulated sites by contrast and regulation direction."}
seq_list <- significant_sites |>
  dplyr::mutate(.grp = paste(contrast, regulation, sep = "_")) |>
  dplyr::group_by(.grp) |>
  dplyr::summarize(
    seqs = list(toupper(SequenceWindow)),
    .groups = "drop"
  ) |>
  with(setNames(seqs, .grp))

if (length(seq_list) > 0) {
  ggseqlogo(
    seq_list,
    ncol = 2,
    seq_type = "aa",
    method = "probability"
  )
} else {
  cat("No significant sites found for sequence logo generation.\n")
}
```

# Difference Logo Analysis

Visualize the difference in amino acid enrichment between upregulated and downregulated sites.

```{r figDifflogo, fig.width=10, fig.height=10, fig.cap="Difference Logos (Upregulated - Downregulated)"}
p_diff <- prophosqua::plot_diff_logo(significant_sites)

if (!is.null(p_diff)) {
  print(p_diff)
} else {
  cat("No contrasts with both upregulated and downregulated sites found for difference logo.\n")
}
```

# Results Summary

```{r summary}
summary_info <- tibble(
  Metric = c("Analysis Type", "FDR Threshold", "FC Threshold",
             "Total Significant Sites", "Upregulated", "Downregulated"),
  Value = c(params$sheet, params$fdr, params$fc,
            nrow(significant_sites),
            sum(significant_sites$regulation == "upregulated"),
            sum(significant_sites$regulation == "downregulated"))
)
knitr::kable(summary_info, caption = "Analysis Summary")
```

# Session Info

```{r sessionInfo}
sessionInfo()
```
