---
title: "PTM Analysis - Sequence Logo"
subtitle: "`r params$sheet` Analysis"
author: "FGCZ"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  xlsx_file: NULL
  sheet: "DPA"
  fdr: 0.05
  fc: 0.6
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    code_folding: hide
    self_contained: true
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Sequence Logo Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

suppressPackageStartupMessages({
  library(dplyr)
  library(ggseqlogo)
  library(readxl)
  library(prophosqua)
})

# Detect pipeline vs vignette mode
pipeline_mode <- !is.null(params$xlsx_file)

# Load shared utilities if in pipeline mode
if (pipeline_mode && file.exists("src/feature_preparation.R")) {
  source("src/feature_preparation.R")
}

```

# Introduction

This document performs sequence logo analysis on significantly regulated sites from **`r params$sheet`** analysis. Sequence motifs help identify active kinases driving PTM changes.

```{r analysisDescription, results='asis'}
desc <- switch(params$sheet,
  "DPA" = "**DPA (Differential PTM Abundance)**: Raw PTM signal changes. Sequence motifs may reflect both abundance and stoichiometry effects.",
  "DPU" = "**DPU (Differential PTM Usage)**: Protein-normalized changes. Sequence motifs reflect genuine kinase activity changes.",
  "CF" = "**CF (CorrectFirst)**: Alternative normalization. Sequence motifs reflect activity changes with different correction approach."
)
cat(desc)
```

# Data Loading

```{r loadData}
if (pipeline_mode) {
  message("Loading data from: ", params$xlsx_file, " (sheet: ", params$sheet, ")")
  data <- readxl::read_xlsx(params$xlsx_file, sheet = params$sheet)
} else {
  # Vignette mode: use example data
  data("combined_test_diff_example", package = "prophosqua")
  data <- combined_test_diff_example
  message("Using example data from prophosqua package")
}

data_info <- tibble(
  Property = c("Mode", "Sheet", "Rows", "Contrasts"),
  Value = c(
    if (pipeline_mode) basename(params$xlsx_file) else "Example data",
    params$sheet,
    nrow(data), paste(unique(data$contrast), collapse = ", ")
  )
)
knitr::kable(
  data_info,
  caption = paste0(
    "Input of this sequence logo report: source workbook, sheet, number of ",
    "site-by-contrast rows read and the contrasts they cover. Every row is one ",
    "phosphosite in one contrast, before any significance filtering."
  )
)
```

# Filter Significant Sites

Filter sites with |log2FC| > `r params$fc` and FDR < `r params$fdr`.

```{r filterSignificant}
# Use shared filtering function with sequence validation
significant_sites <- data |>
  dplyr::filter(!is.na(posInProtein)) |>
  filter_significant_sites(
    fdr_threshold = params$fdr,
    fc_threshold = params$fc,
    require_sequence = TRUE
  )

cat("Found", nrow(significant_sites), "significant sites\n")
```

## Count Significant Sites

```{r tableSignificant}
stopifnot("No significant sites found. Adjust fdr/log2fc thresholds in config." = nrow(significant_sites) > 0)

tx <- as.data.frame(with(significant_sites, table(contrast, regulation, modAA)))
capt <- paste0(
  "Number of significantly regulated sites entering the logos, cross-tabulated by contrast, ",
  "regulation direction and modified residue (", params$sheet, "; |log2FC| > ", params$fc,
  " and FDR < ", params$fdr, "). Freq is the site count of that combination; a direction with ",
  "few sites gives a noisy logo, and S/T/Y counts show which residue class dominates the motif."
)

knitr::kable(tx, caption = capt)
```

## Validate Sequence Window

Ensure the central residue matches the modified amino acid.

```{r validateSeqWindow}
# Use shared validation function
significant_sites <- validate_sequence_window(significant_sites)

cat("After validation:", nrow(significant_sites), "sites remain\n")

# Figure captions, assembled once so they can name the actual thresholds
seqlogo_cap <- paste0(
  "Amino acid composition of the sequence windows around the significantly regulated sites (",
  params$sheet, "), one panel per contrast and regulation direction. The x-axis is the position ",
  "relative to the modified residue in the centre of the window, letter height is the ",
  "probability of that amino acid at that position among the sites of the panel (they sum to 1 ",
  "per position), and letters are coloured by amino acid chemistry. Only sites with |log2FC| > ",
  params$fc, " and FDR < ", params$fdr, " and a validated sequence window are included, so ",
  "panels with few sites show tall letters that are not necessarily meaningful."
)

difflogo_cap <- paste0(
  "Difference in amino acid probabilities between up- and down-regulated sites (",
  params$sheet, "), one panel per contrast that has sites in both directions. The x-axis is ",
  "the position relative to the modified central residue and the y-axis is the probability of ",
  "an amino acid among up-regulated sites minus its probability among down-regulated sites. ",
  "Letters above zero mark residues enriched around up-regulated sites, letters below zero ",
  "residues enriched around down-regulated sites; a position with no directional preference ",
  "stays empty. This contrasts the two motifs directly instead of showing each separately."
)
```

# Sequence Logo Analysis

## Generate Sequence Logos

```{r figSeqlogo, fig.width=9, fig.height=12, fig.cap=seqlogo_cap}
seq_list <- significant_sites |>
  dplyr::mutate(.grp = paste(contrast, regulation, sep = "_")) |>
  dplyr::group_by(.grp) |>
  dplyr::summarize(
    seqs = list(toupper(SequenceWindow)),
    .groups = "drop"
  ) |>
  with(setNames(seqs, .grp))

if (length(seq_list) > 0) {
  ggseqlogo(
    seq_list,
    ncol = 2,
    seq_type = "aa",
    method = "probability"
  )
} else {
  cat("No significant sites found for sequence logo generation.\n")
}
```

# Difference Logo Analysis

Visualize the difference in amino acid enrichment between upregulated and downregulated sites.

```{r figDifflogo, fig.width=10, fig.height=10, fig.cap=difflogo_cap}
p_diff <- prophosqua::plot_diff_logo(significant_sites)

if (!is.null(p_diff)) {
  print(p_diff)
} else {
  cat("No contrasts with both upregulated and downregulated sites found for difference logo.\n")
}
```

# Results Summary

```{r summary}
summary_info <- tibble(
  Metric = c("Analysis Type", "FDR Threshold", "FC Threshold",
             "Total Significant Sites", "Upregulated", "Downregulated"),
  Value = c(params$sheet, params$fdr, params$fc,
            nrow(significant_sites),
            sum(significant_sites$regulation == "upregulated"),
            sum(significant_sites$regulation == "downregulated"))
)
knitr::kable(
  summary_info,
  caption = paste0(
    "Sites behind the logos above: analysis type, the significance thresholds applied, and how ",
    "many sites passed them in total and per direction. Upregulated and Downregulated sum to ",
    "the total; a strong imbalance between them explains an asymmetric difference logo."
  )
)
```

# Session Info

```{r sessionInfo}
sessionInfo()
```
