---
title: "PTM Statistics: DPA, DPU, and CorrectFirst"
format:
  html:
    toc-location: left
    grid:
      sidebar-width: 300px
    include-after-body: fgcz-plot-finder.html
toc: true
params:
  input_h5mu: null
  fdr_threshold: 0.05
  fold_change_threshold: 0.5
vignette: >
  %\VignetteIndexEntry{PTM Statistics: DPA, DPU, and CorrectFirst}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
required_packages <- c(
  "DT",
  "dplyr",
  "fgczQuartoTemplate",
  "ggplot2",
  "ggseqlogo",
  "knitr",
  "prolfqua",
  "prophosqua"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1))
]
if (length(missing_packages)) {
  stop(
    "Missing package(s) required to render this report: ",
    paste(missing_packages, collapse = ", "),
    call. = FALSE
  )
}

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)

input_h5mu <- params$input_h5mu
if (is.null(input_h5mu) || !length(input_h5mu) || !nzchar(input_h5mu)) {
  input_h5mu <- system.file(
    "extdata",
    "ptm_results_example",
    "PTM_results.h5mu",
    package = "prophosqua"
  )
}
if (!nzchar(input_h5mu) || !file.exists(input_h5mu)) {
  stop("PTM statistics MuData file not found: ", input_h5mu, call. = FALSE)
}
input_h5mu <- normalizePath(input_h5mu, mustWork = TRUE)
fdr_threshold <- as.numeric(params$fdr_threshold)
fold_change_threshold <- as.numeric(params$fold_change_threshold)
if (length(fdr_threshold) != 1L || !is.finite(fdr_threshold) ||
    fdr_threshold <= 0 || fdr_threshold > 1) {
  stop("fdr_threshold must be one number in (0, 1].", call. = FALSE)
}
if (length(fold_change_threshold) != 1L ||
    !is.finite(fold_change_threshold) || fold_change_threshold < 0) {
  stop("fold_change_threshold must be one non-negative number.", call. = FALSE)
}
logo_selection <- paste0(
  "FDR < ",
  format(fdr_threshold),
  " and |log2 fold change| > ",
  format(fold_change_threshold)
)

# Production renders use PTM_statistics.h5mu. The bundled vignette example is
# a complete PTM_results.h5mu, which contains the same typed statistics stage.
ptm_stage <- prophosqua::read_ptm_h5mu(input_h5mu)
statistics <- if (inherits(ptm_stage, "PTM_statistics")) {
  ptm_stage
} else if (inherits(ptm_stage, "PTM_results")) {
  ptm_stage$get_statistics()
} else {
  stop("Expected PTM_statistics or PTM_results MuData.", call. = FALSE)
}
inputs <- statistics$get_inputs()
dpa_dpu <- statistics$get_dpa_dpu()
cf <- statistics$get_cf()
# Every analysis is computed from all models; the report shows only the rows
# whose site estimate is observed, and counts the imputed ones.
dpa <- prophosqua::observed_site_estimates(as.data.frame(dpa_dpu$combined_site_prot))
dpu <- prophosqua::observed_site_estimates(as.data.frame(dpa_dpu$combined_test_diff))
# The reported CorrectFirst is corrected with the protein DEA's imputed values.
cf_reported <- statistics$get_cf_reported()
cf_results <- prophosqua::observed_site_estimates(as.data.frame(cf_reported$results), "estimate_type")
estimate_counts <- statistics$get_estimate_counts()
design <- inputs$get_design()
contrasts <- inputs$get_contrasts()
provenance <- inputs$get_provenance()

empty_plot <- function(label) {
  ggplot2::ggplot() +
    ggplot2::annotate("text", x = 0, y = 0, label = label) +
    ggplot2::theme_void()
}

safe_plot <- function(expr, label) {
  tryCatch(expr, error = function(error) {
    empty_plot(paste(label, conditionMessage(error), sep = "\n"))
  })
}

volcano_plot <- function(data, effect, fdr, title) {
  if (!all(c(effect, fdr, "contrast") %in% names(data))) {
    return(empty_plot("The required volcano-plot columns are unavailable."))
  }
  plot_data <- data[
    is.finite(data[[effect]]) & is.finite(data[[fdr]]),
    ,
    drop = FALSE
  ]
  if (!nrow(plot_data)) {
    return(empty_plot("No finite estimates are available."))
  }
  plot_data$significant <-
    plot_data[[fdr]] <= fdr_threshold &
      abs(plot_data[[effect]]) >= fold_change_threshold
  plot_data$negative_log10_fdr <- -log10(pmax(plot_data[[fdr]], .Machine$double.xmin))
  ggplot2::ggplot(
    plot_data,
    ggplot2::aes(
      x = .data[[effect]],
      y = .data$negative_log10_fdr,
      colour = .data$significant
    )
  ) +
    ggplot2::geom_point(alpha = 0.55, size = 1.2) +
    ggplot2::geom_vline(
      xintercept = c(-fold_change_threshold, fold_change_threshold),
      colour = "grey55",
      linetype = "dashed"
    ) +
    ggplot2::geom_hline(
      yintercept = -log10(fdr_threshold),
      colour = "grey55",
      linetype = "dashed"
    ) +
    ggplot2::facet_wrap(ggplot2::vars(.data$contrast), scales = "free") +
    ggplot2::scale_colour_manual(
      values = c(`FALSE` = "grey70", `TRUE` = "#cf3c4f"),
      labels = c(`FALSE` = "No", `TRUE` = "Yes")
    ) +
    ggplot2::labs(
      title = title,
      x = "log2 fold change",
      y = "-log10(FDR)",
      colour = "Passes thresholds"
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(legend.position = "bottom")
}

# A figure faceted by contrast keeps ggplot2's panel grid and gets, per panel,
# the size a single-contrast figure has; its displayed width grows with it, so
# the panels and their text keep their size on the page.
facet_size <- function(data) {
  dims <- ggplot2::wrap_dims(max(1L, dplyr::n_distinct(data$contrast)))
  width <- 1 + 5 * dims[[2]]
  list(
    width = width,
    height = 1.3 + 3.7 * dims[[1]],
    out = paste0(min(100, round(40 * width / 6)), "%")
  )
}

site_protein_plot <- function(data) {
  required <- c("diff.protein", "diff.site", "contrast")
  if (!all(required %in% names(data))) {
    return(empty_plot("Site and protein fold changes are unavailable."))
  }
  plot_data <- data[
    is.finite(data$diff.protein) & is.finite(data$diff.site),
    ,
    drop = FALSE
  ]
  if (!nrow(plot_data)) {
    return(empty_plot("No paired site and protein estimates are available."))
  }
  plot_data$usage_change <- plot_data$diff.site - plot_data$diff.protein
  ggplot2::ggplot(
    plot_data,
    ggplot2::aes(
      x = .data$diff.protein,
      y = .data$diff.site,
      colour = .data$usage_change
    )
  ) +
    ggplot2::geom_abline(slope = 1, intercept = 0, colour = "grey40") +
    ggplot2::geom_point(alpha = 0.55, size = 1.2) +
    ggplot2::facet_wrap(ggplot2::vars(.data$contrast), scales = "free") +
    ggplot2::scale_colour_gradient2(
      low = "#2c7bb6",
      mid = "grey85",
      high = "#d7191c",
      midpoint = 0
    ) +
    ggplot2::labs(
      x = "Protein log2 fold change",
      y = "PTM-site log2 fold change",
      colour = "DPU"
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(legend.position = "bottom")
}

dt_table <- function(data, caption, columns, page_length = 15) {
  columns <- intersect(columns, names(data))
  table <- if (length(columns)) data[, columns, drop = FALSE] else data
  if (!nrow(table)) {
    table <- data.frame(message = "No result rows are available.")
  }
  DT::datatable(
    table,
    caption = caption,
    filter = "bottom",
    rownames = FALSE,
    class = "compact stripe",
    options = list(
      pageLength = page_length,
      scrollX = TRUE,
      scrollY = "420px",
      deferRender = TRUE
    )
  )
}

estimate_count_table <- function(analysis) {
  knitr::kable(dplyr::select(estimate_counts[estimate_counts$analysis == analysis, ], -"analysis"))
}

modification_counts <- function(data) {
  if (!all(c("contrast", "modAA") %in% names(data))) {
    return(data.frame(message = "Modification annotations are unavailable."))
  }
  dplyr::count(data, .data$contrast, .data$modAA, name = "sites") |>
    dplyr::arrange(.data$contrast, dplyr::desc(.data$sites))
}

prepare_logo_sites <- function(data, effect, fdr) {
  required <- c("contrast", "SequenceWindow", "modAA", effect, fdr)
  if (!all(required %in% names(data))) {
    return(data.frame())
  }
  sites <- prophosqua::filter_significant_sites(
    data,
    fdr_col = fdr,
    diff_col = effect,
    fdr_threshold = fdr_threshold,
    fc_threshold = fold_change_threshold,
    require_sequence = TRUE
  )
  sites
}

format_contrast <- function(contrast) {
  gsub("_", " ", contrast, fixed = TRUE)
}

logo_counts <- function(data) {
  if (!nrow(data)) {
    return(data.frame(Message = "No sites pass the displayed thresholds."))
  }
  data |>
    dplyr::count(
      .data$contrast,
      .data$regulation,
      .data$modAA,
      name = "Sites"
    ) |>
    dplyr::transmute(
      Contrast = format_contrast(.data$contrast),
      Direction = dplyr::if_else(
        .data$regulation == "upregulated",
        "Up",
        "Down"
      ),
      `Modified residue` = .data$modAA,
      .data$Sites
    ) |>
    dplyr::arrange(.data$Contrast, .data$Direction, .data$`Modified residue`)
}

logo_theme <- function() {
  ggplot2::theme(
    plot.title = ggplot2::element_blank(),
    strip.text = ggplot2::element_text(size = 10, face = "bold"),
    axis.title = ggplot2::element_text(size = 10),
    axis.text = ggplot2::element_text(size = 8),
    legend.position = "bottom",
    legend.title = ggplot2::element_text(size = 9),
    legend.text = ggplot2::element_text(size = 8),
    panel.spacing = grid::unit(0.8, "lines"),
    plot.margin = ggplot2::margin(4, 8, 4, 8)
  )
}

sequence_logo_plot <- function(data) {
  if (!nrow(data)) {
    return(empty_plot("No sites pass the displayed thresholds."))
  }
  sequence_groups <- data |>
    dplyr::mutate(
      direction = dplyr::if_else(
        .data$regulation == "upregulated",
        "Up",
        "Down"
      )
    ) |>
    dplyr::group_by(.data$contrast, .data$direction) |>
    dplyr::summarize(
      sequences = list(.data$SequenceWindow),
      sites = dplyr::n(),
      .groups = "drop"
    ) |>
    dplyr::mutate(
      group = paste0(
        format_contrast(.data$contrast),
        " · ",
        .data$direction,
        " (n=",
        .data$sites,
        ")"
      )
    ) |>
    dplyr::group_by(.data$group) |>
    dplyr::summarize(
      sequences = .data$sequences,
      .groups = "drop"
    )
  sequence_groups <- stats::setNames(
    sequence_groups$sequences,
    sequence_groups$group
  )
  ggseqlogo::ggseqlogo(
    sequence_groups,
    ncol = 2,
    seq_type = "aa",
    method = "probability"
  ) +
    ggplot2::scale_x_continuous(
      breaks = seq_len(15L),
      labels = -7:7,
      expand = ggplot2::expansion(mult = c(0.01, 0.01))
    ) +
    ggplot2::labs(
      title = NULL,
      x = "Position relative to modified residue",
      y = "Amino-acid probability",
      fill = "Amino-acid chemistry"
    ) +
    logo_theme()
}

difference_logo_plot <- function(data) {
  if (!nrow(data)) {
    return(empty_plot("No sites pass the displayed thresholds."))
  }
  plot <- prophosqua::plot_diff_logo(data)
  if (is.null(plot)) {
    return(empty_plot("Both directions are required for a difference logo."))
  }
  contrasts <- unique(data$contrast)
  labels <- vapply(contrasts, function(contrast) {
    contrast_data <- data[data$contrast == contrast, , drop = FALSE]
    paste0(
      format_contrast(contrast),
      " · Up n=",
      sum(contrast_data$regulation == "upregulated"),
      " − Down n=",
      sum(contrast_data$regulation == "downregulated")
    )
  }, character(1))
  names(labels) <- paste0(contrasts, " (Up-Down)")
  layer_data <- plot$layers[[1L]]$data
  layer_data$seq_group <- unname(labels[layer_data$seq_group])
  plot$layers[[1L]]$data <- layer_data
  plot +
    ggplot2::scale_x_continuous(
      breaks = seq_len(15L),
      labels = -7:7,
      expand = ggplot2::expansion(mult = c(0.01, 0.01))
    ) +
    ggplot2::labs(
      title = NULL,
      x = "Position relative to modified residue",
      y = "Δ probability (up − down)",
      fill = "Amino-acid chemistry"
    ) +
    logo_theme()
}

# Both logo figures lay their panels out two per row, a sequence logo per
# contrast and direction, a difference logo per contrast; the figure grows by
# a fixed height per row, so more contrasts do not squeeze the panels.
logo_height <- function(panels) 1.5 + 1.9 * max(1L, ceiling(panels / 2))
sequence_logo_height <- function(data) logo_height(nrow(dplyr::distinct(data, .data$contrast, .data$regulation)))
difference_logo_height <- function(data) logo_height(dplyr::n_distinct(data$contrast))

dpa_logo_sites <- prepare_logo_sites(dpa, "diff.site", "FDR.site")
dpu_logo_sites <- prepare_logo_sites(dpu, "diff_diff", "FDR_I")
cf_logo_sites <- prepare_logo_sites(cf_results, "diff.site", "FDR.site")

source_paths <- provenance$paths
if (is.null(source_paths)) {
  source_paths <- character()
}
generated_at <- format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")
```

```{r report-metadata}
#| results: asis
cat(sprintf(
  '<div id="fgcz-report-metadata" hidden data-generated-at="%s"></div>',
  generated_at
))
```

::: {.panel-tabset}

# Overview

<img src="`r knitr::image_uri(system.file("report-assets", "ptm-statistics.svg", package = "prophosqua"))`" alt="Visual abstract showing enriched PTM and total-proteome measurements entering DPA, DPU, and CorrectFirst DPU branches, followed by comparable plots and result tables." style="width: 100%; max-height: 22rem; object-fit: contain;">

This report compares three complementary views of PTM regulation from one completed MuData analysis. **DPA** describes the observed change at each PTM site. **DPU** subtracts the protein-level change after modelling. **CorrectFirst DPU** first corrects site abundances by their protein abundances and then fits the model.

```{r}
#| label: tbl-overview-input-summary
#| tbl-cap: "Input summary for the completed PTM statistics analysis."
overview <- data.frame(
  Measure = c("Samples", "PTM sites", "Proteins", "Contrasts"),
  Count = c(
    nrow(design),
    dplyr::n_distinct(dpa_dpu$combined_site_prot$site),
    dplyr::n_distinct(dpa_dpu$combined_site_prot$protein_Id),
    length(contrasts)
  )
)
knitr::kable(overview)
```

::: {layout-ncol="2"}

```{r}
#| label: tbl-overview-contrasts
#| tbl-cap: "Contrasts evaluated by DPA, DPU and CorrectFirst DPU."
knitr::kable(data.frame(Contrast = names(cf$contrasts), Definition = unname(cf$contrasts)))
```

```{r}
#| label: tbl-overview-match-rate
#| tbl-cap: "PTM sites paired with a protein-level estimate in each contrast; only these enter DPU and CorrectFirst DPU."
knitr::kable(dpa_dpu$match_rates, digits = 1)
```

:::

The detailed input references and software versions are recorded in **Session Info**.

# DPA

::: {.panel-tabset}

## Summary

Differential PTM abundance reports the site-level effect before protein correction. It is the direct answer to “did this measured PTM site change?”

::: {layout-ncol="2"}

```{r}
#| label: tbl-dpa-estimate-counts
#| tbl-cap: "DPA site-contrast rows by site estimate type. All are computed; this report shows only the observed ones."
estimate_count_table("DPA")
```

```{r}
#| label: tbl-dpa-modification-counts
#| tbl-cap: "Observed DPA rows by contrast and modified amino acid."
knitr::kable(modification_counts(dpa))
```

:::

## Volcano

```{r}
#| label: fig-dpa-volcano
#| fig-width: !expr facet_size(dpa)$width
#| fig-height: !expr facet_size(dpa)$height
#| out-width: !expr facet_size(dpa)$out
#| fig-cap: >-
#|   Differential PTM abundance by contrast. Each point is one PTM site; the
#|   x-axis is its site-level log2 fold change and the y-axis is -log10(FDR).
#|   Red points pass the displayed absolute fold-change and FDR thresholds.
volcano_plot(dpa, "diff.site", "FDR.site", "Differential PTM abundance")
```

## Site vs protein

```{r}
#| label: fig-dpa-site-protein
#| fig-width: !expr facet_size(dpa)$width
#| fig-height: !expr facet_size(dpa)$height
#| out-width: !expr facet_size(dpa)$out
#| fig-cap: >-
#|   PTM-site log2 fold change as a function of the matched protein log2 fold
#|   change. Each point is one matched site; the diagonal denotes equal site and
#|   protein change, and colour is their difference, the DPU effect.
site_protein_plot(dpa)
```

## Sequence logos

::: {.panel-tabset}

### Plot

The logos use significant, sequence-validated PTM sites and separate each contrast and regulation direction. Panel labels report the number of sites; letter height is the amino-acid probability at that position. **On-render selection:** `r logo_selection`.

```{r}
#| label: fig-dpa-sequence-logos
#| fig-cap: >-
#|   Sequence motifs around differentially abundant PTM sites. Each panel
#|   contains one contrast and regulation direction after applying the displayed
#|   FDR and absolute fold-change thresholds.
#| fig-width: 12
#| fig-height: !expr sequence_logo_height(dpa_logo_sites)
#| out-width: "60%"
sequence_logo_plot(dpa_logo_sites)
```

### Table

```{r}
#| label: tbl-dpa-logo-counts
#| tbl-cap: "DPA sites entering the sequence logos by contrast, direction, and modified residue."
knitr::kable(logo_counts(dpa_logo_sites), align = c("l", "l", "c", "r"))
```

:::

## Difference logos

These logos are recomputed from MuData during rendering using **`r logo_selection`**.

```{r}
#| label: fig-dpa-difference-logos
#| fig-cap: >-
#|   Difference logos for DPA. Letters above zero are enriched around
#|   upregulated sites; letters below zero are enriched around downregulated
#|   sites in the same contrast.
#| fig-width: 12
#| fig-height: !expr difference_logo_height(dpa_logo_sites)
#| out-width: "60%"
difference_logo_plot(dpa_logo_sites)
```

## Results

```{r}
#| label: tbl-dpa-results
dt_table(
  dpa,
  "Differential PTM abundance results",
  c(
    "protein_Id", "site", "gene_name.site", "modAA", "posInProtein",
    "SequenceWindow", "contrast", "diff.site", "FDR.site",
    "diff.protein", "FDR.protein", "estimate_type.site"
  )
)
```

:::

# DPU

::: {.panel-tabset}

## Summary

Differential PTM usage compares the modelled site change with the modelled protein change. A positive value means the site increased more than its protein; a negative value means it increased less or decreased more. Only a site whose protein has a result has a usage change, so DPU holds the matched sites.

::: {layout-ncol="2"}

```{r}
#| label: tbl-dpu-estimate-counts
#| tbl-cap: "DPU site-contrast rows by site estimate type. All are computed; this report shows only the observed ones; their protein estimate may be imputed."
estimate_count_table("DPU")
```

```{r}
#| label: tbl-dpu-modification-counts
#| tbl-cap: "Observed DPU rows by contrast and modified amino acid."
knitr::kable(modification_counts(dpu))
```

:::

## Volcano

```{r}
#| label: fig-dpu-volcano
#| fig-width: !expr facet_size(dpu)$width
#| fig-height: !expr facet_size(dpu)$height
#| out-width: !expr facet_size(dpu)$out
#| fig-cap: >-
#|   Differential PTM usage by contrast. Each point is one PTM site; the x-axis
#|   is the site-minus-protein log2 fold change and the y-axis is -log10(FDR)
#|   after propagated site and protein uncertainty. Red points pass both
#|   displayed thresholds.
volcano_plot(dpu, "diff_diff", "FDR_I", "Differential PTM usage")
```

## Site vs protein

```{r}
#| label: fig-dpu-site-protein
#| fig-width: !expr facet_size(dpu)$width
#| fig-height: !expr facet_size(dpu)$height
#| out-width: !expr facet_size(dpu)$out
#| fig-cap: >-
#|   Site and protein log2 fold changes used to calculate DPU. Points above the
#|   diagonal have positive PTM usage changes and points below it have negative
#|   usage changes; colour shows the signed site-minus-protein difference.
site_protein_plot(dpu)
```

## Sequence logos

::: {.panel-tabset}

### Plot

The DPU logos group sites by the direction of their protein-adjusted change. Panel labels report the number of sites entering each motif. **On-render selection:** `r logo_selection`.

```{r}
#| label: fig-dpu-sequence-logos
#| fig-cap: >-
#|   Sequence motifs around sites with significant differential PTM usage,
#|   separated by contrast and direction.
#| fig-width: 12
#| fig-height: !expr sequence_logo_height(dpu_logo_sites)
#| out-width: "60%"
sequence_logo_plot(dpu_logo_sites)
```

### Table

```{r}
#| label: tbl-dpu-logo-counts
#| tbl-cap: "DPU sites entering the sequence logos by contrast, direction, and modified residue."
knitr::kable(logo_counts(dpu_logo_sites), align = c("l", "l", "c", "r"))
```

:::

## Difference logos

These logos are recomputed from MuData during rendering using **`r logo_selection`**.

```{r}
#| label: fig-dpu-difference-logos
#| fig-cap: >-
#|   Difference logos for DPU, showing the amino-acid probability around
#|   upregulated sites minus that around downregulated sites in each contrast.
#| fig-width: 12
#| fig-height: !expr difference_logo_height(dpu_logo_sites)
#| out-width: "60%"
difference_logo_plot(dpu_logo_sites)
```

## Results

```{r}
#| label: tbl-dpu-results
dt_table(
  dpu,
  "Differential PTM usage results",
  c(
    "protein_Id", "site", "gene_name.site", "modAA", "posInProtein",
    "SequenceWindow", "contrast", "diff.site", "diff.protein", "diff_diff",
    "SE_I", "pValue_I", "FDR_I"
  )
)
```

:::

# CorrectFirst DPU

::: {.panel-tabset}

## Summary

CorrectFirst subtracts the protein abundance from the site abundance on the log2 scale and adds back the sample's median protein abundance before modelling. A protein value missing in a sample is taken from the protein DEA's `lm_impute` model, so every site whose protein was quantified is corrected. The fitted site effect is then the change in PTM usage.

::: {layout-ncol="2"}

```{r}
#| label: tbl-cf-estimate-counts
#| tbl-cap: "CorrectFirst site-contrast rows by estimate type. All are computed; this report shows only the observed ones."
estimate_count_table("CF")
```

```{r}
#| label: tbl-cf-modification-counts
#| tbl-cap: "Observed CorrectFirst rows by contrast and modified amino acid."
knitr::kable(modification_counts(cf_results))
```

:::

## Quality control

The two plots assess the protein-corrected sample-level values entering the CorrectFirst model.

::: {layout-ncol="2"}

```{r}
#| label: fig-cf-density
#| fig-cap: >-
#|   Distribution of protein-corrected log2 PTM-site abundance by sample. Each
#|   curve is one sample; shifts or shape differences indicate global changes in
#|   the corrected values entering the CorrectFirst model.
#| fig-width: 6
#| fig-height: 5
safe_plot(
  cf_reported$ptm_data$get_Plotter()$intensity_distribution_density(),
  "CorrectFirst density plot unavailable."
)
```

```{r}
#| label: fig-cf-pca
#| fig-cap: >-
#|   Principal-component analysis of protein-corrected PTM-site abundances.
#|   Points are samples, coloured by experimental group; the projection uses
#|   imputation only for visualization and does not refit the statistical model.
#| fig-width: 6
#| fig-height: 5
safe_plot(
  prolfqua::AggregateLimpa$new(
    cf_reported$ptm_data,
    impute_only = TRUE
  )$aggregate()$get_Plotter()$pca(),
  "CorrectFirst PCA unavailable."
)
```

:::

## Volcano

```{r}
#| label: fig-cf-volcano
#| fig-width: !expr facet_size(cf_results)$width
#| fig-height: !expr facet_size(cf_results)$height
#| out-width: !expr facet_size(cf_results)$out
#| fig-cap: >-
#|   CorrectFirst differential PTM usage by contrast. Each point is one PTM
#|   site modelled after sample-level protein correction; the x-axis is its log2
#|   fold change and the y-axis is -log10(FDR). Red points pass both displayed
#|   thresholds.
volcano_plot(cf_results, "diff.site", "FDR.site", "CorrectFirst differential PTM usage")
```

## Sequence logos

::: {.panel-tabset}

### Plot

The CorrectFirst logos use the direction of the effect fitted after sample-level protein correction. Panel labels report the number of sites entering each motif. **On-render selection:** `r logo_selection`.

```{r}
#| label: fig-cf-sequence-logos
#| fig-cap: >-
#|   Sequence motifs around sites with significant CorrectFirst differential
#|   PTM usage, separated by contrast and direction.
#| fig-width: 12
#| fig-height: !expr sequence_logo_height(cf_logo_sites)
#| out-width: "60%"
sequence_logo_plot(cf_logo_sites)
```

### Table

```{r}
#| label: tbl-cf-logo-counts
#| tbl-cap: "CorrectFirst sites entering the sequence logos by contrast, direction, and modified residue."
knitr::kable(logo_counts(cf_logo_sites), align = c("l", "l", "c", "r"))
```

:::

## Difference logos

These logos are recomputed from MuData during rendering using **`r logo_selection`**.

```{r}
#| label: fig-cf-difference-logos
#| fig-cap: >-
#|   Difference logos for CorrectFirst DPU, showing the amino-acid probability
#|   around upregulated sites minus that around downregulated sites.
#| fig-width: 12
#| fig-height: !expr difference_logo_height(cf_logo_sites)
#| out-width: "60%"
difference_logo_plot(cf_logo_sites)
```

## Results

```{r}
#| label: tbl-cf-results
dt_table(
  cf_results,
  "CorrectFirst differential PTM usage results",
  c(
    "protein_Id", "site", "gene_name", "modAA", "posInProtein",
    "SequenceWindow", "contrast", "diff.site", "std.error",
    "statistic.site", "p.value", "FDR.site", "estimate_type"
  )
)
```

:::

# Session Info

::: {.panel-tabset}

## Report provenance

```{r}
#| label: tbl-report-provenance
#| tbl-cap: "Report provenance and the final MuData input reference."
qmd_source <- fgczQuartoTemplate::fgcz_qmd_source_link()
provenance_table <- data.frame(
  Field = c(
    "Input MuData",
    "Source DEA artifacts",
    "Contrasts",
    "FDR threshold",
    "Absolute log2 fold-change threshold",
    "Report package",
    "Generated at",
    "Qmd source"
  ),
  Value = c(
    input_h5mu,
    paste(source_paths, collapse = "; "),
    paste(names(contrasts), collapse = "; "),
    format(fdr_threshold),
    format(fold_change_threshold),
    paste0("prophosqua ", utils::packageVersion("prophosqua")),
    generated_at,
    qmd_source
  ),
  stringsAsFactors = FALSE
)
provenance_table <- provenance_table[nzchar(provenance_table$Value), , drop = FALSE]
knitr::kable(provenance_table, escape = FALSE)
```

## R session info

```{r session-info}
sessionInfo()
```

:::

:::
