---
title: "PTM Enrichment: PTM-SEA, Kinase GSEA, and MEA"
format:
  html:
    toc-location: left
    grid:
      sidebar-width: 300px
    include-after-body: fgcz-plot-finder.html
toc: true
params:
  input_h5mu: null
  fdr_threshold: 0.1
  top_terms: 10
  running_terms: 1
vignette: >
  %\VignetteIndexEntry{PTM Enrichment: PTM-SEA, Kinase GSEA, and MEA}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
required_packages <- c(
  "clusterProfiler",
  "DT",
  "dplyr",
  "enrichplot",
  "fgczQuartoTemplate",
  "ggplot2",
  "ggridges",
  "jsonlite",
  "knitr",
  "prophosqua",
  "protsea"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1))
]
if (length(missing_packages)) {
  stop(
    "Missing package(s) required to render this report: ",
    paste(missing_packages, collapse = ", "),
    call. = FALSE
  )
}

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5.5,
  fig.dpi = 110,
  out.width = "88%",
  fig.align = "center"
)

input_h5mu <- params$input_h5mu
if (is.null(input_h5mu) || !length(input_h5mu) || !nzchar(input_h5mu)) {
  input_h5mu <- system.file(
    "extdata",
    "ptm_results_example.h5mu",
    package = "prophosqua"
  )
}
if (!nzchar(input_h5mu) || !file.exists(input_h5mu)) {
  stop("Final PTM_results.h5mu not found: ", input_h5mu, call. = FALSE)
}
input_h5mu <- normalizePath(input_h5mu, mustWork = TRUE)
fdr_threshold <- as.numeric(params$fdr_threshold)
top_terms <- as.integer(params$top_terms)
running_terms <- as.integer(params$running_terms)
if (length(fdr_threshold) != 1L || !is.finite(fdr_threshold) ||
    fdr_threshold <= 0 || fdr_threshold > 1) {
  stop("fdr_threshold must be one number in (0, 1].", call. = FALSE)
}
if (length(top_terms) != 1L || is.na(top_terms) || top_terms < 1L) {
  stop("top_terms must be one positive integer.", call. = FALSE)
}
if (length(running_terms) != 1L || is.na(running_terms) || running_terms < 1L) {
  stop("running_terms must be one positive integer.", call. = FALSE)
}

# The report has one data boundary: the complete typed final MuData object.
ptm_results <- prophosqua::read_ptm_h5mu(
  input_h5mu,
  prophosqua::PTM_results
)
statistics <- ptm_results$get_statistics()
inputs <- statistics$get_inputs()
provenance <- inputs$get_provenance()
source_paths <- provenance$paths
if (is.null(source_paths)) {
  source_paths <- character()
}

analysis_labels <- c(DPA = "DPA", DPU = "DPU", CF = "CorrectFirst DPU")
method_labels <- c(
  PTMSEA = "PTM-SEA",
  KinaseGSEA = "Kinase GSEA",
  MEA = "MEA"
)
category_names <- c(PTMSEA = "PTM-SEA", KinaseGSEA = "KinaseLib", MEA = "MEA")

empty_terms <- function() {
  data.frame(
    analysis = character(),
    method = character(),
    contrast = character(),
    term = character(),
    NES = numeric(),
    FDR = numeric(),
    direction = character(),
    mapped_sites = integer(),
    set_size = integer(),
    leading_edge_sites = character(),
    stringsAsFactors = FALSE
  )
}

scalar_number <- function(value) {
  if (is.null(value) || !length(value)) {
    return(NA_real_)
  }
  as.numeric(unlist(value, use.names = FALSE)[[1L]])
}

term_table_from_payload <- function(payload, analysis, method) {
  category <- category_names[[method]]
  tables <- lapply(names(payload$data), function(contrast_name) {
    terms <- payload$data[[contrast_name]]$categories[[category]]$terms
    if (!length(terms)) {
      return(empty_terms())
    }
    do.call(rbind, lapply(terms, function(term) {
      data.frame(
        analysis = analysis,
        method = method_labels[[method]],
        contrast = contrast_name,
        term = as.character(term$description),
        NES = scalar_number(term$enrichment_score),
        FDR = scalar_number(term$fdr),
        direction = as.character(term$direction),
        mapped_sites = as.integer(scalar_number(term$genes_mapped)),
        set_size = as.integer(scalar_number(term$genes_in_set)),
        leading_edge_sites = paste(
          as.character(unlist(term$leading_edge_ids, use.names = FALSE)),
          collapse = ";"
        ),
        stringsAsFactors = FALSE
      )
    }))
  })
  dplyr::bind_rows(tables)
}

read_method_data <- function(analysis, method) {
  document <- ptm_results$get_enrichment_document(method, analysis)
  payload <- jsonlite::fromJSON(document$json, simplifyVector = FALSE)
  category <- category_names[[method]]
  decoded <- protsea::decode_gsea_json(document$json)
  native <- lapply(decoded, function(contrast) contrast[[category]])
  list(
    analysis = analysis,
    method = method,
    label = method_labels[[method]],
    category = category,
    version = document$version,
    checksum = document$sha256,
    contrasts = names(payload$data),
    table = term_table_from_payload(payload, analysis, method),
    native = native
  )
}

enrichment <- lapply(names(analysis_labels), function(analysis) {
  stats::setNames(
    lapply(names(method_labels), function(method) {
      read_method_data(analysis, method)
    }),
    names(method_labels)
  )
})
names(enrichment) <- names(analysis_labels)

summary_table <- function(data) {
  dplyr::bind_rows(lapply(data$contrasts, function(contrast_name) {
    current <- data$table[data$table$contrast == contrast_name, , drop = FALSE]
    data.frame(
      Contrast = gsub("_", " ", contrast_name, fixed = TRUE),
      `Tested terms` = nrow(current),
      `FDR-significant` = sum(current$FDR <= fdr_threshold, na.rm = TRUE),
      `Positive NES` = sum(
        current$FDR <= fdr_threshold & current$NES > 0,
        na.rm = TRUE
      ),
      `Negative NES` = sum(
        current$FDR <= fdr_threshold & current$NES < 0,
        na.rm = TRUE
      ),
      check.names = FALSE
    )
  }))
}

empty_plot <- function(label) {
  ggplot2::ggplot() +
    ggplot2::annotate("text", x = 0, y = 0, label = label) +
    ggplot2::theme_void()
}

dot_plot <- function(data) {
  if (!nrow(data$table)) {
    return(empty_plot("No enrichment terms are available."))
  }
  if (!is.null(data$native)) {
    usable <- data$native[vapply(data$native, function(x) nrow(x@result) > 0L, logical(1))]
    if (!length(usable)) {
      return(empty_plot("No enrichment terms are available."))
    }
    object <- if (length(usable) == 1L) {
      usable[[1L]]
    } else {
      clusterProfiler::merge_result(usable)
    }
    return(
      enrichplot::dotplot(object, showCategory = top_terms) +
        ggplot2::theme(
          axis.text.y = ggplot2::element_text(size = 8),
          legend.position = "bottom"
        )
    )
  }
  plot_data <- data$table |>
    dplyr::group_by(.data$contrast) |>
    dplyr::slice_min(.data$FDR, n = top_terms, with_ties = FALSE) |>
    dplyr::ungroup() |>
    dplyr::mutate(negative_log10_fdr = -log10(pmax(.data$FDR, 1e-12)))
  ggplot2::ggplot(
    plot_data,
    ggplot2::aes(
      x = .data$NES,
      y = stats::reorder(.data$term, .data$NES),
      colour = .data$NES,
      size = .data$negative_log10_fdr
    )
  ) +
    ggplot2::geom_point() +
    ggplot2::geom_vline(xintercept = 0, colour = "grey60", linetype = "dashed") +
    ggplot2::scale_colour_gradient2(low = "#2b6cb0", mid = "grey85", high = "#c53030") +
    ggplot2::facet_wrap(~ .data$contrast, scales = "free_y") +
    ggplot2::theme_bw() +
    ggplot2::theme(legend.position = "bottom", axis.text.y = ggplot2::element_text(size = 8)) +
    ggplot2::labs(x = "Normalized enrichment score", y = NULL, colour = "NES", size = "-log10(FDR)")
}

heatmap_plot <- function(data) {
  if (!nrow(data$table)) {
    return(empty_plot("No enrichment terms are available."))
  }
  filter_threshold <- if (any(data$table$FDR <= 0.15, na.rm = TRUE)) 0.15 else 1
  prophosqua::plot_enrichment_heatmap(
    data$table,
    item_col = "term",
    fdr_col = "FDR",
    fdr_filter = filter_threshold,
    n_top = top_terms,
    item_label_col = "term"
  ) +
    ggplot2::theme(axis.text.y = ggplot2::element_text(size = 8))
}

volcano_plot <- function(data) {
  finite <- data$table[
    is.finite(data$table$NES) & is.finite(data$table$FDR),
    ,
    drop = FALSE
  ]
  if (!nrow(finite)) {
    return(empty_plot("No finite enrichment statistics are available."))
  }
  prophosqua::plot_enrichment_volcano(
    finite,
    item_col = "term",
    fdr_col = "FDR",
    fdr_threshold = fdr_threshold,
    label_fdr_threshold = fdr_threshold,
    n_labels = min(5L, top_terms)
  ) +
    ggplot2::theme(legend.position = "bottom")
}

running_score_plots <- function(data) {
  if (is.null(data$native)) {
    return(list())
  }
  plots <- list()
  for (contrast_name in names(data$native)) {
    result <- data$native[[contrast_name]]
    if (!nrow(result@result)) {
      next
    }
    top <- result@result |>
      dplyr::arrange(.data$p.adjust, .data$pvalue) |>
      utils::head(running_terms)
    for (i in seq_len(nrow(top))) {
      term_id <- top$ID[[i]]
      title <- paste0(
        gsub("_", " ", contrast_name, fixed = TRUE),
        " · ",
        top$Description[[i]],
        " (NES=", round(top$NES[[i]], 2),
        ", FDR=", signif(top$p.adjust[[i]], 2), ")"
      )
      plots[[paste(contrast_name, term_id, sep = "__")]] <- enrichplot::gseaplot2(
        result,
        geneSetID = term_id,
        title = title,
        pvalue_table = FALSE
      )
    }
  }
  plots
}

print_running_scores <- function(data) {
  plots <- running_score_plots(data)
  if (!length(plots)) {
    cat("::: {.callout-note}\nNo enrichment term is available for a running-score plot.\n:::\n")
    return(invisible(NULL))
  }
  for (plot in plots) {
    print(plot)
  }
  invisible(NULL)
}

nonempty_native_results <- function(data) {
  if (is.null(data$native)) {
    return(list())
  }
  data$native[vapply(
    data$native,
    function(result) nrow(result@result) > 0L,
    logical(1)
  )]
}

combined_native_result <- function(data) {
  usable <- nonempty_native_results(data)
  if (!length(usable)) {
    return(NULL)
  }
  if (length(usable) == 1L) {
    return(usable[[1L]])
  }
  clusterProfiler::merge_result(usable)
}

rank_distribution_plots <- function(data) {
  usable <- nonempty_native_results(data)
  lapply(names(usable), function(contrast_name) {
    result <- usable[[contrast_name]]
    n_show <- min(top_terms, nrow(result@result))
    enrichplot::ridgeplot(
      result,
      showCategory = as.numeric(n_show),
      fill = "p.adjust"
    ) +
      ggplot2::labs(
        title = paste0(
          gsub("_", " ", contrast_name, fixed = TRUE),
          " · ranked-site distributions"
        ),
        x = "Site ranking statistic",
        fill = "FDR"
      ) +
      ggplot2::theme(legend.position = "bottom")
  })
}

print_rank_distributions <- function(data) {
  plots <- rank_distribution_plots(data)
  if (!length(plots)) {
    cat("::: {.callout-note}\nNo enrichment term is available for a rank-distribution plot.\n:::\n")
    return(invisible(NULL))
  }
  for (plot in plots) {
    print(plot)
  }
  invisible(NULL)
}

gene_set_network_plot <- function(data) {
  object <- combined_native_result(data)
  if (is.null(object)) {
    return(empty_plot("No enrichment term is available for a gene-set network."))
  }
  enrichplot::cnetplot(
    object,
    showCategory = min(4L, top_terms),
    node_label = "category",
    cex.params = list(category_label = 0.65)
  ) +
    ggplot2::labs(
      title = paste0(analysis_labels[[data$analysis]], " · ", data$label),
      subtitle = "Leading terms and their shared ranked sites"
    )
}

term_similarity_plots <- function(data) {
  object <- combined_native_result(data)
  if (is.null(object)) {
    return(list())
  }
  result <- if (methods::is(object, "compareClusterResult")) {
    object@compareClusterResult
  } else {
    object@result
  }
  if (length(unique(result$ID)) < 3L) {
    return(list())
  }
  n_show <- min(12L, top_terms, length(unique(result$ID)))
  similarity <- enrichplot::pairwise_termsim(object)
  list(
    enrichplot::emapplot(similarity, showCategory = n_show) +
      ggplot2::labs(
        title = "Enrichment map",
        subtitle = "Terms are connected by ranked-site overlap"
      ),
    enrichplot::treeplot(similarity, showCategory = n_show) +
      ggplot2::labs(
        title = "Term-similarity tree",
        subtitle = "Terms are grouped by ranked-site overlap"
      )
  )
}

print_term_similarity <- function(data) {
  plots <- term_similarity_plots(data)
  if (!length(plots)) {
    cat("::: {.callout-note}\nAt least three enrichment terms are required for term-similarity plots.\n:::\n")
    return(invisible(NULL))
  }
  for (plot in plots) {
    print(plot)
  }
  invisible(NULL)
}

result_table <- function(data) {
  if (!nrow(data$table)) {
    return(knitr::kable(data.frame(Status = "No enrichment terms are available.")))
  }
  display <- data$table |>
    dplyr::transmute(
      Contrast = gsub("_", " ", .data$contrast, fixed = TRUE),
      Term = .data$term,
      NES = round(.data$NES, 3),
      FDR = signif(.data$FDR, 3),
      Direction = .data$direction,
      `Mapped sites` = .data$mapped_sites,
      `Set size` = .data$set_size,
      `Leading-edge sites` = .data$leading_edge_sites
    ) |>
    dplyr::arrange(.data$Contrast, .data$FDR)
  DT::datatable(
    display,
    filter = "top",
    rownames = FALSE,
    class = "compact stripe",
    extensions = "Buttons",
    options = list(
      pageLength = 12,
      scrollX = TRUE,
      scrollY = "420px",
      deferRender = TRUE,
      dom = "Bfrtip",
      buttons = c("copy", "csv", "excel")
    )
  )
}

overview <- dplyr::bind_rows(lapply(names(enrichment), function(analysis) {
  dplyr::bind_rows(lapply(enrichment[[analysis]], function(method) {
    data.frame(
      Analysis = analysis_labels[[analysis]],
      Method = method$label,
      Contrasts = length(method$contrasts),
      Terms = nrow(method$table),
      `FDR-significant` = sum(
        method$table$FDR <= fdr_threshold,
        na.rm = TRUE
      ),
      check.names = FALSE
    )
  }))
}))
generated_at <- format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")
```

```{r report-metadata}
#| results: asis
cat(sprintf(
  '<div id="fgcz-report-metadata" hidden data-generated-at="%s"></div>',
  generated_at
))
```

::: {.panel-tabset}

# Overview

::: {.panel-tabset}

## Summary

<img src="`r knitr::image_uri(system.file("report-assets", "ptm-enrichment.svg", package = "prophosqua"))`" alt="Visual abstract showing DPA, DPU, and CorrectFirst rankings entering PTM-SEA, Kinase GSEA, and MEA, followed by comparable plots and tables decoded from JSON in MuData." style="width: 100%; max-height: 22rem; object-fit: contain;">

This report compares three enrichment methods for each PTM analysis. **PTM-SEA** tests curated PTM signatures, **Kinase GSEA** tests predicted kinase substrate sets with clusterProfiler, and **MEA** reports the native motif enrichment produced by the kinase-library Python workflow. Every value is read from validated JSON stored in the final MuData; the report performs no enrichment computation.

```{r}
#| label: tbl-enrichment-overview
#| tbl-cap: "Stored enrichment documents and result counts by PTM analysis and method."
knitr::kable(overview)
```

## Report provenance

```{r}
#| label: tbl-enrichment-provenance
#| tbl-cap: "Report provenance, JSON parameters, and final MuData input reference."
qmd_source <- fgczQuartoTemplate::fgcz_qmd_source_link()
provenance_table <- data.frame(
  Field = c(
    "Input MuData",
    "Source DEA artifacts",
    "Stored documents",
    "JSON schema versions",
    "FDR threshold",
    "Top terms per summary plot",
    "Running-score terms per contrast",
    "Report package",
    "Generated at",
    "Qmd source"
  ),
  Value = c(
    input_h5mu,
    paste(source_paths, collapse = "; "),
    length(ptm_results$get_enrichment_documents()),
    paste(unique(vapply(
      unlist(enrichment, recursive = FALSE),
      function(x) x$version,
      character(1)
    )), collapse = "; "),
    format(fdr_threshold),
    top_terms,
    running_terms,
    paste0("prophosqua ", utils::packageVersion("prophosqua")),
    generated_at,
    qmd_source
  ),
  stringsAsFactors = FALSE
)
provenance_table <- provenance_table[nzchar(provenance_table$Value), , drop = FALSE]
knitr::kable(provenance_table, escape = FALSE)
```

## R session info

```{r session-info}
sessionInfo()
```

:::

# DPA

::: {.panel-tabset}

## PTM-SEA

::: {.panel-tabset}

### Summary

PTM-SEA tests curated kinase, perturbation, pathway, and disease-associated PTM signatures against the DPA site ranking.

```{r}
#| label: tbl-dpa-ptmsea-summary
#| tbl-cap: "DPA PTM-SEA result counts by contrast and enrichment direction."
knitr::kable(summary_table(enrichment$DPA$PTMSEA))
```

### Dot plot

```{r}
#| label: fig-dpa-ptmsea-dotplot
#| fig-width: 11
#| fig-height: 5.2
#| out-width: "94%"
#| fig-cap: "DPA PTM-SEA terms reconstructed from JSON and shown with clusterProfiler enrichment visualization. Position encodes NES, colour encodes adjusted significance, and panels identify contrasts."
dot_plot(enrichment$DPA$PTMSEA)
```

### Heatmap

```{r}
#| label: fig-dpa-ptmsea-heatmap
#| fig-width: 11
#| fig-height: 5.2
#| out-width: "94%"
#| fig-cap: "Normalized enrichment scores for the leading DPA PTM-SEA signatures across contrasts. Tile colour gives NES and symbols indicate FDR thresholds."
heatmap_plot(enrichment$DPA$PTMSEA)
```

### Volcano

```{r}
#| label: fig-dpa-ptmsea-volcano
#| fig-width: 12
#| fig-height: 4.4
#| out-width: "96%"
#| fig-cap: "DPA PTM-SEA normalized enrichment score as a function of FDR. Each point is one signature and each panel is one contrast."
volcano_plot(enrichment$DPA$PTMSEA)
```

### Running score

```{r}
#| label: fig-dpa-ptmsea-running
#| fig-width: 10.5
#| fig-height: 6.2
#| out-width: "92%"
#| fig-cap: "Running enrichment scores for the leading DPA PTM-SEA signature in each contrast, reconstructed from the ordered ranks, signature membership, and weighting parameters stored in JSON."
#| results: asis
print_running_scores(enrichment$DPA$PTMSEA)
```

### Rank distributions

```{r}
#| label: fig-dpa-ptmsea-ridge
#| fig-width: 11
#| fig-height: 5.4
#| out-width: "94%"
#| fig-cap: "DPA PTM-SEA ranked-site distributions for the leading terms. Each ridge shows the ranking statistics of sites belonging to one enriched set and colour encodes FDR."
#| results: asis
print_rank_distributions(enrichment$DPA$PTMSEA)
```

### Gene-set network

```{r}
#| label: fig-dpa-ptmsea-network
#| fig-width: 11
#| fig-height: 6.2
#| out-width: "94%"
#| fig-cap: "DPA PTM-SEA gene-set network. Term nodes connect to their member sites, exposing shared sites across leading enrichment terms and contrasts."
gene_set_network_plot(enrichment$DPA$PTMSEA)
```

### Term similarity

```{r}
#| label: fig-dpa-ptmsea-similarity
#| fig-width: 11
#| fig-height: 6.2
#| out-width: "94%"
#| fig-cap: "DPA PTM-SEA enrichment map and similarity tree. Terms are related by overlap among their ranked-site members."
#| results: asis
print_term_similarity(enrichment$DPA$PTMSEA)
```

### Results

```{r}
#| label: tbl-dpa-ptmsea-results
#| tbl-cap: "Complete DPA PTM-SEA result table decoded from MuData JSON."
result_table(enrichment$DPA$PTMSEA)
```

:::

## Kinase GSEA

::: {.panel-tabset}

### Summary

Kinase GSEA tests predicted kinase substrate sets against the DPA site ranking and preserves the native clusterProfiler result needed for faithful plots.

```{r}
#| label: tbl-dpa-kinase-summary
#| tbl-cap: "DPA Kinase GSEA result counts by contrast and enrichment direction."
knitr::kable(summary_table(enrichment$DPA$KinaseGSEA))
```

### Dot plot

```{r}
#| label: fig-dpa-kinase-dotplot
#| fig-width: 11
#| fig-height: 5.2
#| out-width: "94%"
#| fig-cap: "DPA Kinase GSEA terms reconstructed from JSON and shown with clusterProfiler enrichment visualization."
dot_plot(enrichment$DPA$KinaseGSEA)
```

### Heatmap

```{r}
#| label: fig-dpa-kinase-heatmap
#| fig-width: 11
#| fig-height: 5.2
#| out-width: "94%"
#| fig-cap: "Normalized enrichment scores for the leading DPA kinase substrate sets across contrasts."
heatmap_plot(enrichment$DPA$KinaseGSEA)
```

### Volcano

```{r}
#| label: fig-dpa-kinase-volcano
#| fig-width: 12
#| fig-height: 4.4
#| out-width: "96%"
#| fig-cap: "DPA Kinase GSEA normalized enrichment score as a function of FDR."
volcano_plot(enrichment$DPA$KinaseGSEA)
```

### Running score

```{r}
#| label: fig-dpa-kinase-running
#| fig-width: 10.5
#| fig-height: 6.2
#| out-width: "92%"
#| fig-cap: "Running enrichment scores for the leading DPA kinase substrate set in each contrast, reconstructed from JSON."
#| results: asis
print_running_scores(enrichment$DPA$KinaseGSEA)
```

### Rank distributions

```{r}
#| label: fig-dpa-kinase-ridge
#| fig-width: 11
#| fig-height: 5.4
#| out-width: "94%"
#| fig-cap: "DPA Kinase GSEA ranked-site distributions for the leading terms. Each ridge shows the ranking statistics of sites belonging to one enriched set and colour encodes FDR."
#| results: asis
print_rank_distributions(enrichment$DPA$KinaseGSEA)
```

### Gene-set network

```{r}
#| label: fig-dpa-kinase-network
#| fig-width: 11
#| fig-height: 6.2
#| out-width: "94%"
#| fig-cap: "DPA Kinase GSEA gene-set network. Term nodes connect to their member sites, exposing shared sites across leading enrichment terms and contrasts."
gene_set_network_plot(enrichment$DPA$KinaseGSEA)
```

### Term similarity

```{r}
#| label: fig-dpa-kinase-similarity
#| fig-width: 11
#| fig-height: 6.2
#| out-width: "94%"
#| fig-cap: "DPA Kinase GSEA enrichment map and similarity tree. Terms are related by overlap among their ranked-site members."
#| results: asis
print_term_similarity(enrichment$DPA$KinaseGSEA)
```

### Results

```{r}
#| label: tbl-dpa-kinase-results
#| tbl-cap: "Complete DPA Kinase GSEA result table decoded from MuData JSON."
result_table(enrichment$DPA$KinaseGSEA)
```

:::

## MEA

::: {.panel-tabset}

### Summary

MEA is the native kinase-library/GSEApy motif enrichment. Its ranks, substrate sets, leading substrates, statistics, source running scores, and hit positions are stored in the shared JSON structure. Protsea constructs a temporary clusterProfiler result from that JSON only for plotting.

```{r}
#| label: tbl-dpa-mea-summary
#| tbl-cap: "DPA MEA result counts by contrast and enrichment direction."
knitr::kable(summary_table(enrichment$DPA$MEA))
```

### Dot plot

```{r}
#| label: fig-dpa-mea-dotplot
#| fig-width: 11
#| fig-height: 5.2
#| out-width: "94%"
#| fig-cap: "DPA MEA motif enrichment. Position and colour encode the native NES and point size encodes FDR."
dot_plot(enrichment$DPA$MEA)
```

### Heatmap

```{r}
#| label: fig-dpa-mea-heatmap
#| fig-width: 11
#| fig-height: 5.2
#| out-width: "94%"
#| fig-cap: "Native MEA normalized enrichment scores for the leading DPA kinase motif sets across contrasts."
heatmap_plot(enrichment$DPA$MEA)
```

### Volcano

```{r}
#| label: fig-dpa-mea-volcano
#| fig-width: 12
#| fig-height: 4.4
#| out-width: "96%"
#| fig-cap: "Native DPA MEA normalized enrichment score as a function of FDR."
volcano_plot(enrichment$DPA$MEA)
```

### Running score

```{r}
#| label: mea-dpa-running-note
#| results: asis
print_running_scores(enrichment$DPA$MEA)
```

### Rank distributions

```{r}
#| label: fig-dpa-mea-ridge
#| fig-width: 11
#| fig-height: 5.4
#| out-width: "94%"
#| fig-cap: "DPA MEA ranked-site distributions for the leading terms. Each ridge shows the ranking statistics of sites belonging to one enriched set and colour encodes FDR."
#| results: asis
print_rank_distributions(enrichment$DPA$MEA)
```

### Gene-set network

```{r}
#| label: fig-dpa-mea-network
#| fig-width: 11
#| fig-height: 6.2
#| out-width: "94%"
#| fig-cap: "DPA MEA gene-set network. Term nodes connect to their member sites, exposing shared sites across leading enrichment terms and contrasts."
gene_set_network_plot(enrichment$DPA$MEA)
```

### Term similarity

```{r}
#| label: fig-dpa-mea-similarity
#| fig-width: 11
#| fig-height: 6.2
#| out-width: "94%"
#| fig-cap: "DPA MEA enrichment map and similarity tree. Terms are related by overlap among their ranked-site members."
#| results: asis
print_term_similarity(enrichment$DPA$MEA)
```

### Results

```{r}
#| label: tbl-dpa-mea-results
#| tbl-cap: "Complete native DPA MEA result table decoded from MuData JSON."
result_table(enrichment$DPA$MEA)
```

:::

:::

# DPU

::: {.panel-tabset}

## PTM-SEA

::: {.panel-tabset}

### Summary

PTM-SEA tests curated PTM signatures against the DPU ranking of site change relative to protein change.

```{r}
#| label: tbl-dpu-ptmsea-summary
#| tbl-cap: "DPU PTM-SEA result counts by contrast and enrichment direction."
knitr::kable(summary_table(enrichment$DPU$PTMSEA))
```

### Dot plot

```{r}
#| label: fig-dpu-ptmsea-dotplot
#| fig-width: 11
#| fig-height: 5.2
#| out-width: "94%"
#| fig-cap: "DPU PTM-SEA terms reconstructed from JSON and shown with clusterProfiler enrichment visualization."
dot_plot(enrichment$DPU$PTMSEA)
```

### Heatmap

```{r}
#| label: fig-dpu-ptmsea-heatmap
#| fig-width: 11
#| fig-height: 5.2
#| out-width: "94%"
#| fig-cap: "Normalized enrichment scores for the leading DPU PTM-SEA signatures across contrasts."
heatmap_plot(enrichment$DPU$PTMSEA)
```

### Volcano

```{r}
#| label: fig-dpu-ptmsea-volcano
#| fig-width: 12
#| fig-height: 4.4
#| out-width: "96%"
#| fig-cap: "DPU PTM-SEA normalized enrichment score as a function of FDR."
volcano_plot(enrichment$DPU$PTMSEA)
```

### Running score

```{r}
#| label: fig-dpu-ptmsea-running
#| fig-width: 10.5
#| fig-height: 6.2
#| out-width: "92%"
#| fig-cap: "Running enrichment scores for the leading DPU PTM-SEA signature in each contrast, reconstructed from JSON."
#| results: asis
print_running_scores(enrichment$DPU$PTMSEA)
```

### Rank distributions

```{r}
#| label: fig-dpu-ptmsea-ridge
#| fig-width: 11
#| fig-height: 5.4
#| out-width: "94%"
#| fig-cap: "DPU PTM-SEA ranked-site distributions for the leading terms. Each ridge shows the ranking statistics of sites belonging to one enriched set and colour encodes FDR."
#| results: asis
print_rank_distributions(enrichment$DPU$PTMSEA)
```

### Gene-set network

```{r}
#| label: fig-dpu-ptmsea-network
#| fig-width: 11
#| fig-height: 6.2
#| out-width: "94%"
#| fig-cap: "DPU PTM-SEA gene-set network. Term nodes connect to their member sites, exposing shared sites across leading enrichment terms and contrasts."
gene_set_network_plot(enrichment$DPU$PTMSEA)
```

### Term similarity

```{r}
#| label: fig-dpu-ptmsea-similarity
#| fig-width: 11
#| fig-height: 6.2
#| out-width: "94%"
#| fig-cap: "DPU PTM-SEA enrichment map and similarity tree. Terms are related by overlap among their ranked-site members."
#| results: asis
print_term_similarity(enrichment$DPU$PTMSEA)
```

### Results

```{r}
#| label: tbl-dpu-ptmsea-results
#| tbl-cap: "Complete DPU PTM-SEA result table decoded from MuData JSON."
result_table(enrichment$DPU$PTMSEA)
```

:::

## Kinase GSEA

::: {.panel-tabset}

### Summary

Kinase GSEA tests predicted kinase substrate sets against the DPU ranking.

```{r}
#| label: tbl-dpu-kinase-summary
#| tbl-cap: "DPU Kinase GSEA result counts by contrast and enrichment direction."
knitr::kable(summary_table(enrichment$DPU$KinaseGSEA))
```

### Dot plot

```{r}
#| label: fig-dpu-kinase-dotplot
#| fig-width: 11
#| fig-height: 5.2
#| out-width: "94%"
#| fig-cap: "DPU Kinase GSEA terms reconstructed from JSON and shown with clusterProfiler enrichment visualization."
dot_plot(enrichment$DPU$KinaseGSEA)
```

### Heatmap

```{r}
#| label: fig-dpu-kinase-heatmap
#| fig-width: 11
#| fig-height: 5.2
#| out-width: "94%"
#| fig-cap: "Normalized enrichment scores for the leading DPU kinase substrate sets across contrasts."
heatmap_plot(enrichment$DPU$KinaseGSEA)
```

### Volcano

```{r}
#| label: fig-dpu-kinase-volcano
#| fig-width: 12
#| fig-height: 4.4
#| out-width: "96%"
#| fig-cap: "DPU Kinase GSEA normalized enrichment score as a function of FDR."
volcano_plot(enrichment$DPU$KinaseGSEA)
```

### Running score

```{r}
#| label: fig-dpu-kinase-running
#| fig-width: 10.5
#| fig-height: 6.2
#| out-width: "92%"
#| fig-cap: "Running enrichment scores for the leading DPU kinase substrate set in each contrast, reconstructed from JSON."
#| results: asis
print_running_scores(enrichment$DPU$KinaseGSEA)
```

### Rank distributions

```{r}
#| label: fig-dpu-kinase-ridge
#| fig-width: 11
#| fig-height: 5.4
#| out-width: "94%"
#| fig-cap: "DPU Kinase GSEA ranked-site distributions for the leading terms. Each ridge shows the ranking statistics of sites belonging to one enriched set and colour encodes FDR."
#| results: asis
print_rank_distributions(enrichment$DPU$KinaseGSEA)
```

### Gene-set network

```{r}
#| label: fig-dpu-kinase-network
#| fig-width: 11
#| fig-height: 6.2
#| out-width: "94%"
#| fig-cap: "DPU Kinase GSEA gene-set network. Term nodes connect to their member sites, exposing shared sites across leading enrichment terms and contrasts."
gene_set_network_plot(enrichment$DPU$KinaseGSEA)
```

### Term similarity

```{r}
#| label: fig-dpu-kinase-similarity
#| fig-width: 11
#| fig-height: 6.2
#| out-width: "94%"
#| fig-cap: "DPU Kinase GSEA enrichment map and similarity tree. Terms are related by overlap among their ranked-site members."
#| results: asis
print_term_similarity(enrichment$DPU$KinaseGSEA)
```

### Results

```{r}
#| label: tbl-dpu-kinase-results
#| tbl-cap: "Complete DPU Kinase GSEA result table decoded from MuData JSON."
result_table(enrichment$DPU$KinaseGSEA)
```

:::

## MEA

::: {.panel-tabset}

### Summary

MEA reports the native kinase-library motif enrichment over the DPU ranking.

```{r}
#| label: tbl-dpu-mea-summary
#| tbl-cap: "DPU MEA result counts by contrast and enrichment direction."
knitr::kable(summary_table(enrichment$DPU$MEA))
```

### Dot plot

```{r}
#| label: fig-dpu-mea-dotplot
#| fig-width: 11
#| fig-height: 5.2
#| out-width: "94%"
#| fig-cap: "DPU MEA motif enrichment using the native NES and FDR stored in JSON."
dot_plot(enrichment$DPU$MEA)
```

### Heatmap

```{r}
#| label: fig-dpu-mea-heatmap
#| fig-width: 11
#| fig-height: 5.2
#| out-width: "94%"
#| fig-cap: "Native MEA normalized enrichment scores for the leading DPU kinase motif sets across contrasts."
heatmap_plot(enrichment$DPU$MEA)
```

### Volcano

```{r}
#| label: fig-dpu-mea-volcano
#| fig-width: 12
#| fig-height: 4.4
#| out-width: "96%"
#| fig-cap: "Native DPU MEA normalized enrichment score as a function of FDR."
volcano_plot(enrichment$DPU$MEA)
```

### Running score

```{r}
#| label: mea-dpu-running-note
#| results: asis
print_running_scores(enrichment$DPU$MEA)
```

### Rank distributions

```{r}
#| label: fig-dpu-mea-ridge
#| fig-width: 11
#| fig-height: 5.4
#| out-width: "94%"
#| fig-cap: "DPU MEA ranked-site distributions for the leading terms. Each ridge shows the ranking statistics of sites belonging to one enriched set and colour encodes FDR."
#| results: asis
print_rank_distributions(enrichment$DPU$MEA)
```

### Gene-set network

```{r}
#| label: fig-dpu-mea-network
#| fig-width: 11
#| fig-height: 6.2
#| out-width: "94%"
#| fig-cap: "DPU MEA gene-set network. Term nodes connect to their member sites, exposing shared sites across leading enrichment terms and contrasts."
gene_set_network_plot(enrichment$DPU$MEA)
```

### Term similarity

```{r}
#| label: fig-dpu-mea-similarity
#| fig-width: 11
#| fig-height: 6.2
#| out-width: "94%"
#| fig-cap: "DPU MEA enrichment map and similarity tree. Terms are related by overlap among their ranked-site members."
#| results: asis
print_term_similarity(enrichment$DPU$MEA)
```

### Results

```{r}
#| label: tbl-dpu-mea-results
#| tbl-cap: "Complete native DPU MEA result table decoded from MuData JSON."
result_table(enrichment$DPU$MEA)
```

:::

:::

# CorrectFirst DPU

::: {.panel-tabset}

## PTM-SEA

::: {.panel-tabset}

### Summary

PTM-SEA tests curated PTM signatures against the ranking fitted after sample-level protein correction.

```{r}
#| label: tbl-cf-ptmsea-summary
#| tbl-cap: "CorrectFirst PTM-SEA result counts by contrast and enrichment direction."
knitr::kable(summary_table(enrichment$CF$PTMSEA))
```

### Dot plot

```{r}
#| label: fig-cf-ptmsea-dotplot
#| fig-width: 11
#| fig-height: 5.2
#| out-width: "94%"
#| fig-cap: "CorrectFirst PTM-SEA terms reconstructed from JSON and shown with clusterProfiler enrichment visualization."
dot_plot(enrichment$CF$PTMSEA)
```

### Heatmap

```{r}
#| label: fig-cf-ptmsea-heatmap
#| fig-width: 11
#| fig-height: 5.2
#| out-width: "94%"
#| fig-cap: "Normalized enrichment scores for the leading CorrectFirst PTM-SEA signatures across contrasts."
heatmap_plot(enrichment$CF$PTMSEA)
```

### Volcano

```{r}
#| label: fig-cf-ptmsea-volcano
#| fig-width: 12
#| fig-height: 4.4
#| out-width: "96%"
#| fig-cap: "CorrectFirst PTM-SEA normalized enrichment score as a function of FDR."
volcano_plot(enrichment$CF$PTMSEA)
```

### Running score

```{r}
#| label: fig-cf-ptmsea-running
#| fig-width: 10.5
#| fig-height: 6.2
#| out-width: "92%"
#| fig-cap: "Running enrichment scores for the leading CorrectFirst PTM-SEA signature in each contrast, reconstructed from JSON."
#| results: asis
print_running_scores(enrichment$CF$PTMSEA)
```

### Rank distributions

```{r}
#| label: fig-cf-ptmsea-ridge
#| fig-width: 11
#| fig-height: 5.4
#| out-width: "94%"
#| fig-cap: "CorrectFirst PTM-SEA ranked-site distributions for the leading terms. Each ridge shows the ranking statistics of sites belonging to one enriched set and colour encodes FDR."
#| results: asis
print_rank_distributions(enrichment$CF$PTMSEA)
```

### Gene-set network

```{r}
#| label: fig-cf-ptmsea-network
#| fig-width: 11
#| fig-height: 6.2
#| out-width: "94%"
#| fig-cap: "CorrectFirst PTM-SEA gene-set network. Term nodes connect to their member sites, exposing shared sites across leading enrichment terms and contrasts."
gene_set_network_plot(enrichment$CF$PTMSEA)
```

### Term similarity

```{r}
#| label: fig-cf-ptmsea-similarity
#| fig-width: 11
#| fig-height: 6.2
#| out-width: "94%"
#| fig-cap: "CorrectFirst PTM-SEA enrichment map and similarity tree. Terms are related by overlap among their ranked-site members."
#| results: asis
print_term_similarity(enrichment$CF$PTMSEA)
```

### Results

```{r}
#| label: tbl-cf-ptmsea-results
#| tbl-cap: "Complete CorrectFirst PTM-SEA result table decoded from MuData JSON."
result_table(enrichment$CF$PTMSEA)
```

:::

## Kinase GSEA

::: {.panel-tabset}

### Summary

Kinase GSEA tests predicted kinase substrate sets against the CorrectFirst ranking.

```{r}
#| label: tbl-cf-kinase-summary
#| tbl-cap: "CorrectFirst Kinase GSEA result counts by contrast and enrichment direction."
knitr::kable(summary_table(enrichment$CF$KinaseGSEA))
```

### Dot plot

```{r}
#| label: fig-cf-kinase-dotplot
#| fig-width: 11
#| fig-height: 5.2
#| out-width: "94%"
#| fig-cap: "CorrectFirst Kinase GSEA terms reconstructed from JSON and shown with clusterProfiler enrichment visualization."
dot_plot(enrichment$CF$KinaseGSEA)
```

### Heatmap

```{r}
#| label: fig-cf-kinase-heatmap
#| fig-width: 11
#| fig-height: 5.2
#| out-width: "94%"
#| fig-cap: "Normalized enrichment scores for the leading CorrectFirst kinase substrate sets across contrasts."
heatmap_plot(enrichment$CF$KinaseGSEA)
```

### Volcano

```{r}
#| label: fig-cf-kinase-volcano
#| fig-width: 12
#| fig-height: 4.4
#| out-width: "96%"
#| fig-cap: "CorrectFirst Kinase GSEA normalized enrichment score as a function of FDR."
volcano_plot(enrichment$CF$KinaseGSEA)
```

### Running score

```{r}
#| label: fig-cf-kinase-running
#| fig-width: 10.5
#| fig-height: 6.2
#| out-width: "92%"
#| fig-cap: "Running enrichment scores for the leading CorrectFirst kinase substrate set in each contrast, reconstructed from JSON."
#| results: asis
print_running_scores(enrichment$CF$KinaseGSEA)
```

### Rank distributions

```{r}
#| label: fig-cf-kinase-ridge
#| fig-width: 11
#| fig-height: 5.4
#| out-width: "94%"
#| fig-cap: "CorrectFirst Kinase GSEA ranked-site distributions for the leading terms. Each ridge shows the ranking statistics of sites belonging to one enriched set and colour encodes FDR."
#| results: asis
print_rank_distributions(enrichment$CF$KinaseGSEA)
```

### Gene-set network

```{r}
#| label: fig-cf-kinase-network
#| fig-width: 11
#| fig-height: 6.2
#| out-width: "94%"
#| fig-cap: "CorrectFirst Kinase GSEA gene-set network. Term nodes connect to their member sites, exposing shared sites across leading enrichment terms and contrasts."
gene_set_network_plot(enrichment$CF$KinaseGSEA)
```

### Term similarity

```{r}
#| label: fig-cf-kinase-similarity
#| fig-width: 11
#| fig-height: 6.2
#| out-width: "94%"
#| fig-cap: "CorrectFirst Kinase GSEA enrichment map and similarity tree. Terms are related by overlap among their ranked-site members."
#| results: asis
print_term_similarity(enrichment$CF$KinaseGSEA)
```

### Results

```{r}
#| label: tbl-cf-kinase-results
#| tbl-cap: "Complete CorrectFirst Kinase GSEA result table decoded from MuData JSON."
result_table(enrichment$CF$KinaseGSEA)
```

:::

## MEA

::: {.panel-tabset}

### Summary

MEA reports the native kinase-library motif enrichment over the CorrectFirst ranking.

```{r}
#| label: tbl-cf-mea-summary
#| tbl-cap: "CorrectFirst MEA result counts by contrast and enrichment direction."
knitr::kable(summary_table(enrichment$CF$MEA))
```

### Dot plot

```{r}
#| label: fig-cf-mea-dotplot
#| fig-width: 11
#| fig-height: 5.2
#| out-width: "94%"
#| fig-cap: "CorrectFirst MEA motif enrichment using the native NES and FDR stored in JSON."
dot_plot(enrichment$CF$MEA)
```

### Heatmap

```{r}
#| label: fig-cf-mea-heatmap
#| fig-width: 11
#| fig-height: 5.2
#| out-width: "94%"
#| fig-cap: "Native MEA normalized enrichment scores for the leading CorrectFirst kinase motif sets across contrasts."
heatmap_plot(enrichment$CF$MEA)
```

### Volcano

```{r}
#| label: fig-cf-mea-volcano
#| fig-width: 12
#| fig-height: 4.4
#| out-width: "96%"
#| fig-cap: "Native CorrectFirst MEA normalized enrichment score as a function of FDR."
volcano_plot(enrichment$CF$MEA)
```

### Running score

```{r}
#| label: mea-cf-running-note
#| results: asis
print_running_scores(enrichment$CF$MEA)
```

### Rank distributions

```{r}
#| label: fig-cf-mea-ridge
#| fig-width: 11
#| fig-height: 5.4
#| out-width: "94%"
#| fig-cap: "CorrectFirst MEA ranked-site distributions for the leading terms. Each ridge shows the ranking statistics of sites belonging to one enriched set and colour encodes FDR."
#| results: asis
print_rank_distributions(enrichment$CF$MEA)
```

### Gene-set network

```{r}
#| label: fig-cf-mea-network
#| fig-width: 11
#| fig-height: 6.2
#| out-width: "94%"
#| fig-cap: "CorrectFirst MEA gene-set network. Term nodes connect to their member sites, exposing shared sites across leading enrichment terms and contrasts."
gene_set_network_plot(enrichment$CF$MEA)
```

### Term similarity

```{r}
#| label: fig-cf-mea-similarity
#| fig-width: 11
#| fig-height: 6.2
#| out-width: "94%"
#| fig-cap: "CorrectFirst MEA enrichment map and similarity tree. Terms are related by overlap among their ranked-site members."
#| results: asis
print_term_similarity(enrichment$CF$MEA)
```

### Results

```{r}
#| label: tbl-cf-mea-results
#| tbl-cap: "Complete native CorrectFirst MEA result table decoded from MuData JSON."
result_table(enrichment$CF$MEA)
```

:::

:::

:::
