---
title: "PTM Enrichment: PTM-SEA, Kinase GSEA, and MEA"
format:
  html:
    toc-location: left
    grid:
      sidebar-width: 300px
    include-after-body: fgcz-plot-finder.html
toc: true
params:
  input_h5mu: null
  fdr_threshold: 0.1
  top_terms: 10
  analysis: DPA
vignette: >
  %\VignetteIndexEntry{PTM Enrichment: PTM-SEA, Kinase GSEA, and MEA}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
required_packages <- c(
  "clusterProfiler",
  "DT",
  "dplyr",
  "enrichplot",
  "fgczQuartoTemplate",
  "ggplot2",
  "ggridges",
  "jsonlite",
  "knitr",
  "prophosqua",
  "protsea"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1))
]
if (length(missing_packages)) {
  stop(
    "Missing package(s) required to render this report: ",
    paste(missing_packages, collapse = ", "),
    call. = FALSE
  )
}

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)

input_h5mu <- params$input_h5mu
if (is.null(input_h5mu) || !length(input_h5mu) || !nzchar(input_h5mu)) {
  input_h5mu <- system.file(
    "extdata",
    "ptm_results_example.h5mu",
    package = "prophosqua"
  )
}
if (!nzchar(input_h5mu) || !file.exists(input_h5mu)) {
  stop("Final PTM_results.h5mu not found: ", input_h5mu, call. = FALSE)
}
input_h5mu <- normalizePath(input_h5mu, mustWork = TRUE)
fdr_threshold <- as.numeric(params$fdr_threshold)
top_terms <- as.integer(params$top_terms)
analysis <- toupper(as.character(params$analysis))
if (length(fdr_threshold) != 1L || !is.finite(fdr_threshold) ||
    fdr_threshold <= 0 || fdr_threshold > 1) {
  stop("fdr_threshold must be one number in (0, 1].", call. = FALSE)
}
if (length(top_terms) != 1L || is.na(top_terms) || top_terms < 1L) {
  stop("top_terms must be one positive integer.", call. = FALSE)
}
if (length(analysis) != 1L || !analysis %in% c("DPA", "DPU", "CF")) {
  stop("analysis must be DPA, DPU, or CF.", call. = FALSE)
}

# The report has one data boundary: the complete typed final MuData object.
ptm_results <- prophosqua::read_ptm_h5mu(
  input_h5mu,
  prophosqua::PTM_results
)
statistics <- ptm_results$get_statistics()
inputs <- statistics$get_inputs()
provenance <- inputs$get_provenance()
source_paths <- provenance$paths
if (is.null(source_paths)) {
  source_paths <- character()
}

analysis_labels <- c(DPA = "DPA", DPU = "DPU", CF = "CorrectFirst DPU")
ranking_labels <- c(DPA = "site abundance ranking", DPU = "site usage ranking", CF = "protein-corrected ranking")
method_labels <- c(
  PTMSEA = "PTM-SEA",
  KinaseGSEA = "Kinase GSEA",
  MEA = "MEA"
)
category_names <- c(PTMSEA = "PTM-SEA", KinaseGSEA = "KinaseLib", MEA = "MEA")

empty_terms <- function() {
  data.frame(
    analysis = character(),
    method = character(),
    contrast = character(),
    term = character(),
    NES = numeric(),
    FDR = numeric(),
    direction = character(),
    mapped_sites = integer(),
    set_size = integer(),
    stringsAsFactors = FALSE
  )
}

scalar_number <- function(value) {
  if (is.null(value) || !length(value)) {
    return(NA_real_)
  }
  as.numeric(unlist(value, use.names = FALSE)[[1L]])
}

term_table_from_payload <- function(payload, analysis, method) {
  category <- category_names[[method]]
  tables <- lapply(names(payload$data), function(contrast_name) {
    terms <- payload$data[[contrast_name]]$categories[[category]]$terms
    if (!length(terms)) {
      return(empty_terms())
    }
    do.call(rbind, lapply(terms, function(term) {
      data.frame(
        analysis = analysis,
        method = method_labels[[method]],
        contrast = contrast_name,
        term = as.character(term$description),
        NES = scalar_number(term$enrichment_score),
        FDR = scalar_number(term$fdr),
        direction = as.character(term$direction),
        mapped_sites = as.integer(scalar_number(term$genes_mapped)),
        set_size = as.integer(scalar_number(term$genes_in_set)),
        stringsAsFactors = FALSE
      )
    }))
  })
  dplyr::bind_rows(tables)
}

read_method_data <- function(analysis, method) {
  document <- ptm_results$get_enrichment_document(method, analysis)
  payload <- jsonlite::fromJSON(document$json, simplifyVector = FALSE)
  category <- category_names[[method]]
  decoded <- protsea::decode_gsea_json(document$json)
  native <- lapply(decoded, function(contrast) contrast[[category]])
  list(
    analysis = analysis,
    method = method,
    label = method_labels[[method]],
    category = category,
    version = document$version,
    checksum = document$sha256,
    contrasts = names(payload$data),
    table = term_table_from_payload(payload, analysis, method),
    native = native
  )
}

enrichment <- stats::setNames(
  lapply(names(method_labels), function(method) read_method_data(analysis, method)),
  names(method_labels)
)

summary_table <- function(data) {
  dplyr::bind_rows(lapply(data$contrasts, function(contrast_name) {
    current <- data$table[data$table$contrast == contrast_name, , drop = FALSE]
    data.frame(
      Contrast = gsub("_", " ", contrast_name, fixed = TRUE),
      `Tested terms` = nrow(current),
      `FDR-significant` = sum(current$FDR <= fdr_threshold, na.rm = TRUE),
      `Positive NES` = sum(
        current$FDR <= fdr_threshold & current$NES > 0,
        na.rm = TRUE
      ),
      `Negative NES` = sum(
        current$FDR <= fdr_threshold & current$NES < 0,
        na.rm = TRUE
      ),
      check.names = FALSE
    )
  }))
}

LABEL_WRAP_WIDTH <- 22L

wrap_term_label <- function(labels, width = LABEL_WRAP_WIDTH) {
  vapply(labels, function(label) {
    pieces <- regmatches(label, gregexpr("[^_/ ]+[_/ ]?", label))[[1]]
    lines <- character()
    current <- ""
    for (piece in pieces) {
      if (nzchar(current) && nchar(current) + nchar(piece) > width) {
        lines <- c(lines, current)
        current <- piece
      } else {
        current <- paste0(current, piece)
      }
    }
    paste(c(lines, current), collapse = "\n")
  }, character(1), USE.NAMES = FALSE)
}

wrap_term_labels <- function(object) {
  if (methods::is(object, "compareClusterResult")) {
    object@compareClusterResult$Description <- wrap_term_label(object@compareClusterResult$Description)
  } else {
    object@result$Description <- wrap_term_label(object@result$Description)
  }
  object
}

# enrichplot draws an in-panel pie-size legend for multi-contrast results with no
# switch to turn it off; its layers are the only ones carrying arc data of their own.
drop_pie_legend <- function(plot) {
  is_legend <- vapply(plot$layers, function(layer) {
    is.data.frame(layer$data) && all(c("r", "start", "end", "maxr") %in% names(layer$data))
  }, logical(1))
  plot$layers <- plot$layers[!is_legend]
  plot
}

graph_label_room <- function() {
  list(
    ggplot2::scale_x_continuous(expand = ggplot2::expansion(mult = 0.3)),
    ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = 0.15)),
    ggplot2::theme(
      legend.position = "bottom",
      plot.title.position = "plot",
      plot.margin = ggplot2::margin(8, 12, 8, 12)
    )
  )
}

empty_plot <- function(label) {
  ggplot2::ggplot() +
    ggplot2::annotate("text", x = 0, y = 0, label = label) +
    ggplot2::theme_void()
}

dot_plot <- function(data) {
  if (!nrow(data$table)) {
    return(empty_plot("No enrichment terms are available."))
  }
  if (!is.null(data$native)) {
    usable <- data$native[vapply(data$native, function(x) nrow(x@result) > 0L, logical(1))]
    if (!length(usable)) {
      return(empty_plot("No enrichment terms are available."))
    }
    object <- if (length(usable) == 1L) {
      usable[[1L]]
    } else {
      clusterProfiler::merge_result(usable)
    }
    return(
      enrichplot::dotplot(object, showCategory = top_terms) +
        ggplot2::theme(
          axis.text.y = ggplot2::element_text(size = 8),
          legend.position = "bottom"
        )
    )
  }
  plot_data <- data$table |>
    dplyr::group_by(.data$contrast) |>
    dplyr::slice_min(.data$FDR, n = top_terms, with_ties = FALSE) |>
    dplyr::ungroup() |>
    dplyr::mutate(negative_log10_fdr = -log10(pmax(.data$FDR, 1e-12)))
  ggplot2::ggplot(
    plot_data,
    ggplot2::aes(
      x = .data$NES,
      y = stats::reorder(.data$term, .data$NES),
      colour = .data$NES,
      size = .data$negative_log10_fdr
    )
  ) +
    ggplot2::geom_point() +
    ggplot2::geom_vline(xintercept = 0, colour = "grey60", linetype = "dashed") +
    ggplot2::scale_colour_gradient2(low = "#2b6cb0", mid = "grey85", high = "#c53030") +
    ggplot2::facet_wrap(~ .data$contrast, scales = "free_y") +
    ggplot2::theme_bw() +
    ggplot2::theme(legend.position = "bottom", axis.text.y = ggplot2::element_text(size = 8)) +
    ggplot2::labs(x = "Normalized enrichment score", y = NULL, colour = "NES", size = "-log10(FDR)")
}

heatmap_plot <- function(data) {
  if (!nrow(data$table)) {
    return(empty_plot("No enrichment terms are available."))
  }
  filter_threshold <- if (any(data$table$FDR <= 0.15, na.rm = TRUE)) 0.15 else 1
  prophosqua::plot_enrichment_heatmap(
    data$table,
    item_col = "term",
    fdr_col = "FDR",
    fdr_filter = filter_threshold,
    n_top = top_terms,
    item_label_col = "term"
  ) +
    ggplot2::theme(axis.text.y = ggplot2::element_text(size = 8))
}

volcano_plot <- function(data) {
  finite <- data$table[
    is.finite(data$table$NES) & is.finite(data$table$FDR),
    ,
    drop = FALSE
  ]
  if (!nrow(finite)) {
    return(empty_plot("No finite enrichment statistics are available."))
  }
  prophosqua::plot_enrichment_volcano(
    finite,
    item_col = "term",
    fdr_col = "FDR",
    fdr_threshold = fdr_threshold,
    label_fdr_threshold = fdr_threshold,
    n_labels = min(5L, top_terms)
  ) +
    ggplot2::theme(legend.position = "bottom")
}

nonempty_native_results <- function(data) {
  if (is.null(data$native)) {
    return(list())
  }
  data$native[vapply(
    data$native,
    function(result) nrow(result@result) > 0L,
    logical(1)
  )]
}

combined_native_result <- function(data) {
  usable <- nonempty_native_results(data)
  if (!length(usable)) {
    return(NULL)
  }
  if (length(usable) == 1L) {
    return(usable[[1L]])
  }
  clusterProfiler::merge_result(usable)
}

rank_distribution_plots <- function(data) {
  usable <- nonempty_native_results(data)
  lapply(names(usable), function(contrast_name) {
    result <- usable[[contrast_name]]
    n_show <- min(top_terms, nrow(result@result))
    enrichplot::ridgeplot(
      result,
      showCategory = as.numeric(n_show),
      fill = "p.adjust"
    ) +
      ggplot2::labs(
        title = paste0(
          gsub("_", " ", contrast_name, fixed = TRUE),
          " · ranked-site distributions"
        ),
        x = "Site ranking statistic",
        fill = "FDR"
      ) +
      ggplot2::theme(legend.position = "bottom")
  })
}

print_rank_distributions <- function(data) {
  plots <- rank_distribution_plots(data)
  if (!length(plots)) {
    plots <- list(empty_plot("No enrichment term is available for a rank-distribution plot."))
  }
  for (plot in plots) {
    print(plot)
  }
  invisible(NULL)
}

gene_set_network_plot <- function(data) {
  object <- combined_native_result(data)
  if (is.null(object)) {
    return(empty_plot("No enrichment term is available for a gene-set network."))
  }
  network <- enrichplot::cnetplot(
    wrap_term_labels(object),
    showCategory = min(4L, top_terms),
    node_label = "category"
  )
  drop_pie_legend(network) +
    ggplot2::labs(
      title = paste0(analysis_labels[[data$analysis]], " · ", data$label),
      subtitle = "Leading terms and their shared ranked sites"
    ) +
    graph_label_room()
}

term_similarity_plots <- function(data) {
  object <- combined_native_result(data)
  if (is.null(object)) {
    return(list())
  }
  result <- if (methods::is(object, "compareClusterResult")) {
    object@compareClusterResult
  } else {
    object@result
  }
  if (length(unique(result$ID)) < 3L) {
    return(list())
  }
  n_show <- min(12L, top_terms, length(unique(result$ID)))
  similarity <- enrichplot::pairwise_termsim(object)
  map <- tryCatch(
    drop_pie_legend(enrichplot::emapplot(similarity, showCategory = n_show, label_format = wrap_term_label)) +
      ggplot2::labs(
        title = "Enrichment map",
        subtitle = "Terms are connected by ranked-site overlap"
      ) +
      graph_label_room(),
    error = function(e) NULL
  )
  tree <- tryCatch(
    enrichplot::treeplot(similarity, showCategory = n_show, label_format = wrap_term_label) +
      ggplot2::labs(
        title = "Term-similarity tree",
        subtitle = "Terms are grouped by ranked-site overlap"
      ),
    error = function(e) NULL
  )
  plots <- Filter(Negate(is.null), list(map, tree))
  attr(plots, "tree_unavailable") <- is.null(tree) && !is.null(map)
  plots
}

print_term_similarity <- function(data) {
  plots <- term_similarity_plots(data)
  if (!length(plots)) {
    plots <- list(empty_plot("No term-similarity plot can be drawn for these results."))
  } else if (isTRUE(attr(plots, "tree_unavailable"))) {
    plots <- c(plots, list(empty_plot("A similarity tree could not be constructed for these sets.")))
  }
  for (plot in plots) {
    print(plot)
  }
  invisible(NULL)
}

result_table <- function(data) {
  if (!nrow(data$table)) {
    return(knitr::kable(data.frame(Status = "No enrichment terms are available.")))
  }
  display <- data$table |>
    dplyr::transmute(
      Contrast = gsub("_", " ", .data$contrast, fixed = TRUE),
      Term = .data$term,
      NES = round(.data$NES, 3),
      FDR = signif(.data$FDR, 3),
      Direction = .data$direction,
      `Mapped sites` = .data$mapped_sites,
      `Set size` = .data$set_size
    ) |>
    dplyr::arrange(.data$Contrast, .data$FDR)
  DT::datatable(
    display,
    filter = "top",
    rownames = FALSE,
    class = "compact stripe",
    extensions = "Buttons",
    options = list(
      pageLength = 12,
      scrollX = TRUE,
      scrollY = "420px",
      deferRender = TRUE,
      dom = "Bfrtip",
      buttons = c("copy", "csv", "excel")
    )
  )
}

overview <- dplyr::bind_rows(lapply(enrichment, function(method) {
  data.frame(
    Analysis = analysis_labels[[analysis]],
    Method = method$label,
    Contrasts = length(method$contrasts),
    Terms = nrow(method$table),
    `FDR-significant` = sum(method$table$FDR <= fdr_threshold, na.rm = TRUE),
    check.names = FALSE
  )
}))
generated_at <- format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")
```

```{r report-metadata}
#| results: asis
cat(sprintf(
  '<div id="fgcz-report-metadata" hidden data-generated-at="%s"></div>',
  generated_at
))
```

::: {.panel-tabset}

# Overview

<p class="lead">This <strong>`r analysis_labels[[analysis]]`</strong> report compares three enrichment methods from validated JSON stored in the final MuData.</p>

```{r overview-graphic-data}
#| include: false
graphic <- paste(readLines(system.file("report-assets", "ptm-enrichment.svg", package = "prophosqua"), warn = FALSE), collapse = "\n")
graphic <- gsub("{{ANALYSIS}}", analysis_labels[[analysis]], graphic, fixed = TRUE)
graphic <- gsub("{{RANKING}}", ranking_labels[[analysis]], graphic, fixed = TRUE)
overview_svg <- tempfile(fileext = ".svg")
writeLines(graphic, overview_svg)
```

<img src="`r knitr::image_uri(overview_svg)`" alt="Graphical overview: the `r analysis_labels[[analysis]]` ranking feeds PTM-SEA, Kinase GSEA, and MEA; their stored JSON produces plots and tested-term tables." style="width: 100%; max-height: 19rem; object-fit: contain;">

**PTM-SEA** tests curated PTM signatures, **Kinase GSEA** tests predicted kinase substrate sets, and **MEA** shows native motif enrichment from the kinase-library workflow. Rendering does not rerun enrichment.

```{r}
#| label: tbl-enrichment-overview
#| tbl-cap: "Stored enrichment documents and result counts for this PTM analysis."
knitr::kable(overview)
```

```{r method-helper}
#| include: false
render_method <- function(method) {
  method_data <- enrichment[[method]]
  method_label <- method_labels[[method]]
  id <- paste(tolower(analysis), tolower(method), sep = "-")
  caption <- function(what) sprintf("%s %s: %s", analysis_labels[[analysis]], method_label, what)
  child <- c(
    "::: {.panel-tabset}", "", "## Summary", "",
    "```{r}", sprintf("#| label: tbl-%s-summary", id),
    sprintf("#| tbl-cap: '%s'", caption("result counts by contrast and direction.")),
    "knitr::kable(summary_table(method_data))", "```", "",
    "## Dot plot", "",
    "```{r}", sprintf("#| label: fig-%s-dotplot", id),
    sprintf("#| fig-cap: '%s'", caption("leading enrichment terms reconstructed from JSON.")),
    "dot_plot(method_data)", "```", "",
    "## Heatmap", "",
    "```{r}", sprintf("#| label: fig-%s-heatmap", id),
    sprintf("#| fig-cap: '%s'", caption("normalized enrichment scores across contrasts.")),
    "heatmap_plot(method_data)", "```", "",
    "## Volcano", "",
    "```{r}", sprintf("#| label: fig-%s-volcano", id),
    sprintf("#| fig-cap: '%s'", caption("normalized enrichment score versus FDR.")),
    "volcano_plot(method_data)", "```", "",
    "## Rank distributions", "",
    "```{r}", sprintf("#| label: fig-%s-ridge", id),
    "#| layout-ncol: 2", "#| out-width: '100%'",
    sprintf("#| fig-cap: '%s'", caption("ranked-site distributions of leading terms, one panel per contrast.")),
    "print_rank_distributions(method_data)", "```", "",
    "## Gene-set network", "",
    "```{r}", sprintf("#| label: fig-%s-network", id),
    sprintf("#| fig-cap: '%s'", caption("leading terms and shared sites.")),
    "gene_set_network_plot(method_data)", "```", "",
    "## Term similarity", "",
    "```{r}", sprintf("#| label: fig-%s-similarity", id),
    "#| layout-ncol: 2", "#| out-width: '100%'",
    sprintf("#| fig-cap: '%s'", caption("term overlap map and similarity tree.")),
    "print_term_similarity(method_data)", "```", "",
    "## Results", "",
    "```{r}", sprintf("#| label: tbl-%s-results", id),
    sprintf("#| tbl-cap: '%s'", caption("tested-term statistics decoded from MuData JSON; leading-edge site lists remain in JSON.")),
    "result_table(method_data)", "```", "", ":::"
  )
  knitr::knit_child(text = child, envir = environment(), quiet = TRUE)
}
```

# PTM-SEA

```{r ptmsea-tabs}
#| results: asis
cat(render_method("PTMSEA"), sep = "\n")
```

# Kinase GSEA

```{r kinase-tabs}
#| results: asis
cat(render_method("KinaseGSEA"), sep = "\n")
```

# MEA

```{r mea-tabs}
#| results: asis
cat(render_method("MEA"), sep = "\n")
```

# About methods

## PTM-SEA

PTM-SEA tests curated PTMsigDB phosphosite signatures against the ranked site statistics. Its table retains every tested signature, including those above the report's FDR threshold.

## Kinase GSEA

Kinase GSEA tests substrate sets assigned by the kinase-library motif scan against the same ranked site statistics. It reports tested kinase sets and their adjusted p-values, including non-significant sets.

## MEA

MEA uses the kinase-library workflow's GSEApy motif-enrichment calculation. Its scores and adjusted p-values come from that Python computation; this report reads the stored JSON without recomputing them.

# Session Info

::: {.panel-tabset}

## Report provenance

```{r}
#| label: tbl-enrichment-provenance
#| tbl-cap: "Report provenance and JSON parameters."
provenance_table <- data.frame(
  Field = c(
    "Analysis", "Input MuData", "Source DEA artifacts",
    "Method documents", "JSON schema versions", "FDR threshold",
    "Top terms per summary plot", "Report package", "Generated at", "QMD source"
  ),
  Value = c(
    analysis_labels[[analysis]], input_h5mu, paste(source_paths, collapse = "; "),
    length(enrichment),
    paste(unique(vapply(enrichment, function(x) x$version, character(1))), collapse = "; "),
    format(fdr_threshold), top_terms,
    paste0("prophosqua ", utils::packageVersion("prophosqua")),
    generated_at, fgczQuartoTemplate::fgcz_qmd_source_link()
  ),
  stringsAsFactors = FALSE
)
provenance_table <- provenance_table[nzchar(provenance_table$Value), , drop = FALSE]
knitr::kable(provenance_table, escape = FALSE)
```

## R session info

```{r session-info}
sessionInfo()
```

:::

:::
