---
title: "Integration of PTM Features and Total Proteome Data for Comprehensive Phosphoproteomics Analysis"
author: "Functional Genomics Center Zurich"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
  bookdown::html_document2:
    toc: true
  pdf_document: 
    toc: true
header-includes: 
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhead[CO,CE]{PTM Features and Total Proteome Integration Analysis}
  \fancyfoot[CO,CE]{\textbf{FGCZ} - www.fgcz.ch - 2018}
  \fancyfoot[LE,RO]{\thepage} 
params:
  data: NULL
  grp: NULL
vignette: >
  %\VignetteIndexEntry{FGCZ Two-Group Analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: bibliography2025.bib
editor_options: 
  chunk_output_type: console
always_allow_html: true
---



```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)

grp2 <- params$grp
data_diff_diff <- params$data

# Colours for the per-site estimate type. The keys must match the values found in
# estimate_type.site: a value that is missing from the palette is drawn invisible.
estimate_type_palette <- c(
  observed = "black",
  lod_imputed = "orange",
  missing_fallback = "blue"
)
estimate_type_levels <- names(estimate_type_palette)

# One panel per contrast, at most four panels per row.
facet_nrow <- ceiling(length(unique(data_diff_diff$contrast)) / 4)
fig_height <- max(3.5, 3.4 * facet_nrow)
```

# Overview

This report will be stored at a later point in the LIMS system [_bfabric_](https://fgcz-bfabric.uzh.ch) [@Trker2010] in the 
__project__: `r grp2$Bfabric$projectID`, __order__: `r grp2$project_spec$orderID`, `r if(!is.null(grp2$project_spec$workunitID)){ "with the workunit ID :" }`  `r grp2$project_spec$workunitID`. Currently it is only uploaded as generic zip resource and attached to the result of the order.

In this analysis, both enriched (phosphopeptide enrichment) and non-enriched (total proteome) samples are analyzed separately using standard proteomics software. For the total proteome analysis, protein quantification is performed by aggregating peptide-level measurements to protein-level abundances. For the phosphoproteome analysis, phosphorylation site-specific quantification is performed using PTM-centric workflows. Both datasets are statistically evaluated individually to generate separate reports for differential protein expression (DPE) and differential PTM expression.

After individual analysis, the two datasets are integrated to perform differential PTM usage (DPU) analysis. For all phosphorylation sites where the corresponding protein is also quantified in the total proteome measurement, the protein-level statistics are used to normalize the observed changes in phosphorylation site abundance. This normalization follows the approach described for MSstatsPTM [@Kohler2023yo], where the difference in phosphorylation site levels between conditions is adjusted by the difference in total protein levels for the same comparison. This integrated analysis reveals whether changes in phosphorylation are due to changes in protein abundance or to changes in site-specific modification usage, providing deeper insights into cellular signaling and regulation.

# Numbers

```{r nummms, eval=TRUE, fig.height=5, fig.width=8}
# pick your contrast and subset as before
coi         <- unique(data_diff_diff$contrast)[1]
data_diff_1C <- subset(data_diff_diff, contrast == coi)
tbl_modAA <- table(data_diff_1C$modAA, useNA = "no")

n_sites <- length(unique(data_diff_1C$site))

tbl_combined <- c(tbl_modAA, Total = n_sites)

df_summary <- tibble::enframe(tbl_combined, name = "Modification", value = "Count")
knitr::kable(
  df_summary,
  caption = paste0(
    "Phosphosites quantified in contrast ", coi, ", counted per modified residue. ",
    "Modification is the modified amino acid (S, T or Y; NotLoc marks sites whose position ",
    "within the peptide could not be localized) and Count the number of distinct sites, with ",
    "Total giving the distinct sites over all residue types. Serine dominance is the expected ",
    "pattern in a phosphoproteome."
  )
)

```


# Joining Phosphorylation Sites with Total Proteome

```{r foundInGlobal, eval=TRUE}
tt <- data_diff_diff |>
  dplyr::select(protein_Id, measured_In) |>
  dplyr::distinct() |>
  dplyr::pull(measured_In) |>
  table()

# Calculate percentages
percFoundInGlobal <- tt["both"] / (tt["both"] + tt["site"]) * 100
totProtFromSites <- (tt["both"] + tt["site"])
sitesM <- data_diff_diff |>  
  dplyr::select(protein_Id, site, measured_In) |> 
  dplyr::distinct() |> 
  dplyr::pull(measured_In) |> 
  table()
```

When phosphopeptide enrichment is performed, some peptides are identified for which there is no corresponding protein identification in the total proteome measurement. Therefore, protein-level information may be missing for some phosphorylation sites. Here, for `r percFoundInGlobal`% of all identified phosphoproteins (`r totProtFromSites` total), the corresponding protein is also quantified in the total proteome measurements. For `r sitesM["both"]` phosphorylation sites mapping to a total of `r tt["both"]` proteins, we can perform integrated analysis combining both datasets.


# Integration with Total Proteome

Next, we integrate the phosphorylation site statistics with the protein-level changes using the MSstatsPTM approach [@Kohler2023yo]. We refer to this as *MSstatsPTM adjustment* and refer to these statistics in our outputs as *diff_diff* and *FDR_I*. This analysis reveals differential PTM usage (DPU), which distinguishes between changes due to protein abundance and changes due to site-specific modification usage.

In brief, the difference in phosphorylation site levels between two conditions is normalized by the difference in total protein levels for the same contrast. By doing so, protein-level changes are subtracted from phosphorylation site changes, and the adjusted statistics are used to calculate raw and adjusted p-values (FDR_I) for the normalized difference (diff_diff). This approach reveals whether phosphorylation changes are due to changes in protein abundance or to changes in site-specific modification usage.


```{r tableAllProt}
bb <- data_diff_diff |>
  dplyr::select(all_of(c("site", "protein_Id", "contrast", 
                         "estimate_type.site", "modAA", "diff_diff", "FDR_I")))

DT::datatable(
  bb, 
  filter = "bottom",
  extensions = "Scroller",
  style = "auto",
  class = "compact",
  options = list(deferRender = TRUE,
                 scrollY = 300,
                 scrollX = 400,
                 scroller = TRUE),
  
  caption = paste0(
    "Differential PTM usage for every phosphosite and contrast, after MSstatsPTM adjustment. ",
    "diff_diff is the site log2 fold change minus the log2 fold change of its protein, so it ",
    "measures usage rather than abundance, and FDR_I is its adjusted p-value; ",
    "estimate_type.site states whether the site estimate is observed or was filled in by ",
    "limit-of-detection imputation, and modAA gives the modified residue."
  )) |>
  DT::formatRound(
    columns = c('diff_diff', 'FDR_I'), 
    digits = 3
  )

```


(ref:scatterplot) Phosphosite log2 fold change as a function of the log2 fold change of its protein, one panel per contrast. Each point is one phosphosite: the x-axis is the protein-level difference between the two groups, the y-axis the site-level difference, and the colour marks whether the site estimate is observed (black) or was filled in by limit-of-detection imputation (orange). The dashed blue diagonal marks equal change: points on it change as much as their protein and therefore reflect protein abundance, while points off it indicate changed modification usage. Sites without a quantified protein are absent.

```{r scatterplot, fig.cap="(ref:scatterplot)", fig.width=9, fig.height=fig_height, include = TRUE, eval = TRUE}

datax <- data_diff_diff |>
  dplyr::filter(!is.na(diff.protein), !is.na(diff.site)) |>
  # draw the few imputed sites last so they are not buried under the observed ones
  dplyr::arrange(factor(estimate_type.site, levels = estimate_type_levels))

ggplot2::ggplot(datax, ggplot2::aes(x = diff.protein, y = diff.site,
                                    colour = estimate_type.site)) +
  ggplot2::geom_point(size = 0.5, alpha = 0.3) +
  ggplot2::geom_abline(slope = 1, intercept = 0,
                       linetype = "dashed", colour = "steelblue") +
  ggplot2::facet_wrap(~contrast, nrow = facet_nrow,
                      labeller = ggplot2::label_wrap_gen(width = 24)) +
  ggplot2::scale_colour_manual(values = estimate_type_palette, name = "site estimate") +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::theme(strip.text = ggplot2::element_text(size = 7),
                 legend.position = "bottom") +
  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 2, alpha = 1)))

```


(ref:volcanoplot) Differential PTM usage per contrast, one panel per contrast. Each point is one phosphosite: the x-axis is the protein-adjusted difference diff_diff (site log2 fold change minus protein log2 fold change) and the y-axis the negative log10 of the adjusted p-value FDR_I, so points far left or right are used less or more than their protein would explain and points high up are the confident ones. The red dotted horizontal line marks FDR = `r grp2$processing_options$FDR_threshold` and the green dotted vertical lines the difference thresholds of minus and plus `r grp2$processing_options$diff_threshold`; orange points were filled in by limit-of-detection imputation and deserve more caution than the black ones.


```{r volcanoplot, fig.cap = "(ref:volcanoplot)", fig.width=9, fig.height=fig_height, include = TRUE, eval = TRUE}

datax <- data_diff_diff |>
  dplyr::filter(!is.na(diff_diff), !is.na(FDR_I)) |>
  # draw the few imputed sites last so they are not buried under the observed ones
  dplyr::arrange(factor(estimate_type.site, levels = estimate_type_levels))

ggplot2::ggplot(datax, ggplot2::aes(x = diff_diff, y = -log10(FDR_I),
                                    colour = estimate_type.site)) +
  ggplot2::geom_point(size = 0.5, alpha = 0.3) +
  ggplot2::geom_vline(xintercept = c(-grp2$processing_options$diff_threshold,
                                     grp2$processing_options$diff_threshold),
                      linetype = "dotted", colour = "green4") +
  ggplot2::geom_hline(yintercept = -log10(grp2$processing_options$FDR_threshold),
                      linetype = "dotted", colour = "red") +
  ggplot2::facet_wrap(~contrast, nrow = facet_nrow,
                      labeller = ggplot2::label_wrap_gen(width = 24)) +
  ggplot2::scale_colour_manual(values = estimate_type_palette, name = "site estimate") +
  ggplot2::labs(y = expression(-log[10](FDR_I))) +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::theme(strip.text = ggplot2::element_text(size = 7),
                 legend.position = "bottom") +
  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 2, alpha = 1)))

```

# References
