---
title: "Kinase Activity (Kinase Library + GSEA)"
subtitle: "`r params$analysis_type`"
author: "FGCZ"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  objects_rds: NULL
  analysis_type: "DPA"
  output_dir: NULL
  top_genesets: 3
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    code_folding: hide
    self_contained: true
vignette: >
  %\VignetteIndexEntry{Kinase Activity (Kinase Library + GSEA)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 10
)

# Detect pipeline vs vignette mode
pipeline_mode <- !is.null(params$objects_rds)
```

## Introduction

This report demonstrates kinase activity inference using the **Kinase Library** motif predictions. Unlike PTMsigDB (which uses curated kinase-substrate relationships), Kinase Library predicts kinase-substrate assignments based on position-specific scoring matrices (PSSMs) derived from peptide library screens.

We use the `term2gene.csv` file generated by the `scan-motifs` CLI, which contains kinase-sequence assignments.

**Approach:** GSEA (Gene Set Enrichment Analysis) using ranked phosphosite statistics.

## Analysis Details

| Parameter | Value |
|-----------|-------|
| Analysis Type | `r params$analysis_type` |
| Mode | `r if(pipeline_mode) "Pipeline" else "Vignette"` |

## Load Data and Kinase Library Predictions

```{r load-packages}
library(clusterProfiler)
library(dplyr)
library(tidyr)
library(readxl)
library(ggplot2)
library(enrichplot)
library(purrr)
library(forcats)
library(writexl)
library(patchwork)
library(DT)
library(prophosqua)
```

```{r load-data}
if (pipeline_mode) {
  kl <- readRDS(params$objects_rds)
} else {
  # Standalone mode: compute from the example data shipped with the package,
  # through the same function the pipeline calls.
  kl <- prophosqua:::compute_kinaselib_gsea_example()
}

# The settings the computation actually ran with, so that the captions below
# quote the analysis rather than a second copy of its configuration. `params` is
# a locked binding during a render and cannot be extended, so they get their own
# name; `params` keeps only what the caller passed.
inputs <- kl$analysis_inputs

output_dir <- if (!is.null(params$output_dir)) {
  params$output_dir
} else {
  dirname(inputs$xlsx_file)
}

knitr::kable(
  kl$data_info,
  caption = paste0(
    "Input of this kinase-library GSEA report: source workbook, sheet, table dimensions and ",
    "the contrasts covered. Every row is one phosphosite in one contrast, before ranking."
  )
)
```

```{r load-kinase-library}
term2gene <- kl$term2gene
term2gene_df <- kl$term2gene_df

knitr::kable(
  kl$kl_info,
  caption = paste0(
    "Kinase-substrate predictions used as gene sets: the term2gene file produced by the ",
    "kinase-library motif scan, with the number of kinases and of predicted kinase-site ",
    "assignments it contains. A site is assigned to a kinase when its sequence window scores ",
    "above the configured motif percentile, so one site can belong to several kinases."
  )
)
```

## Kinase-Substrate Assignment Statistics

```{r assignment-stats}
knitr::kable(
  kl$assignment_stats,
  caption = paste0(
    "How much of the measured phosphoproteome carries a kinase motif assignment. The first ",
    "row counts the distinct sequence windows in the differential analysis, the second the ",
    "windows the motif scan assigned to at least one kinase, the third those present in both ",
    "and therefore usable for GSEA, and Coverage is that intersection as a percentage of the ",
    "measured windows. Low coverage limits how many kinase sets can be tested."
  )
)
```

```{r kinase-stats}
knitr::kable(
  kl$kinase_stats,
  caption = paste0(
    "Size of the kinase substrate sets that GSEA tests. The first rows describe the predicted ",
    "set sizes over all kinases (count, mean, median and range of substrates per kinase), the ",
    "last two how many of those substrates are actually measured here on average and how many ",
    "kinases keep at least 15 measured substrates. Sets that stay below the min_size ",
    "threshold are dropped before testing."
  )
)
```

## Prepare Ranked Lists

```{r prepare-ranks}
knitr::kable(
  kl$ranks_info,
  caption = paste0(
    "Ranked site list submitted to GSEA per contrast (", params$analysis_type,
    "): Sites is the number of ranked sequence windows in that contrast's input vector, ",
    "ordered by ", inputs$stat_col, " from most up- to most down-regulated. Contrasts differ ",
    "when sites are missing values in one comparison but not another."
  )
)
```

## GSEA Analysis

```{r run-gsea}
gsea_results <- kl$gsea_results
gsea_info <- kl$gsea_info
has_gsea_results <- kl$has_results

knitr::kable(
  gsea_info,
  caption = paste0(
    "GSEA outcome per contrast (", params$analysis_type, "): number of kinase substrate sets ",
    "reaching FDR < 0.25, in either direction, out of the sets that passed the size filter. ",
    "A contrast with zero hits points at low motif coverage or a weak overall response."
  )
)
```

```{r noGseaResults, eval=!has_gsea_results, results='asis'}
cat("\n\n# No GSEA Results\n\n")
cat("No kinase gene sets passed the size filter (min_size=", inputs$min_size, ").\n")
cat("This typically means too few phosphosites overlap with kinase library assignments.\n\n")
cat("**Coverage was:", kl$n_overlap_seqs, "sites out of", kl$n_our_sequences, "**\n\n")
```

```{r prepareCaptions, eval=has_gsea_results}
# Extract all results into data frame using shared function
all_clean <- extract_gsea_results(gsea_results) |>
  mutate(kinase = ID)

# Contrast order and figure captions, assembled once for the loops below
contrast_ids <- unique(all_clean$contrast)

dotplot_caps <- paste0(
  "Kinase enrichment in contrast ", contrast_ids, " (", params$analysis_type,
  "): normalized enrichment score of the 30 kinases with the smallest FDR. Each point is one ",
  "kinase substrate set predicted by the kinase-library motif scan; the x-axis is the NES from ",
  "GSEA over sites ranked by ", inputs$stat_col, ", so NES > 0 means the predicted substrates ",
  "accumulate among up-regulated sites and NES < 0 among down-regulated ones. Point size is ",
  "-log10(FDR), point colour follows the NES (blue negative, red positive), faded points are ",
  "kinases with FDR >= 0.1, and the dashed line marks NES = 0."
)

merged_dotplot_cap <- paste0(
  "The 15 most enriched kinase substrate sets compared across all contrasts (",
  params$analysis_type, ", FDR < 0.25). Rows are kinases, columns are contrasts, the x-axis ",
  "position within a column is the fraction of the kinase's substrates found in the leading ",
  "edge, point size the number of those substrates and point colour the adjusted p-value. ",
  "Gaps mark kinases that missed the cutoff in that contrast, which makes contrast-specific ",
  "kinase responses visible."
)

individual_dotplot_cap <- paste0(
  "Per-contrast kinase enrichment as clusterProfiler dotplots (", params$analysis_type,
  "), one panel per contrast, two panels per row. Within a panel each row is one of the 15 ",
  "most enriched kinases, the x-axis is the fraction of its substrates in the leading edge, ",
  "point size the number of those substrates and point colour the adjusted p-value. A panel ",
  "reading \"No significant kinases\" means no set passed the cutoff in that contrast."
)

heatmap_cap <- paste0(
  "Normalized enrichment score per kinase and contrast (", params$analysis_type, "). Rows are ",
  "the 30 kinases with the smallest FDR in any contrast (only kinases reaching FDR < 0.25 ",
  "somewhere are eligible), ordered by NES; columns are the contrasts. Tile colour is the NES ",
  "(blue: predicted substrates enriched among down-regulated sites, red: among up-regulated ",
  "sites, white: no enrichment) and asterisks give the FDR of that kinase in that contrast ",
  "(* < 0.1, ** < 0.05, *** < 0.01). Empty tiles mean the kinase was not tested there."
)

volcano_cap <- paste0(
  "Enrichment strength versus significance for every tested kinase substrate set, one panel ",
  "per contrast (", params$analysis_type, "). Each point is one kinase; the x-axis is the NES ",
  "(negative: substrates enriched among down-regulated sites, positive: among up-regulated ",
  "sites) and the y-axis -log10(FDR), with the horizontal dashed line at FDR = 0.1 and the ",
  "vertical dashed line at NES = 0. Colour encodes the direction, faded points are kinases ",
  "above the FDR cutoff, and the five kinases with the smallest FDR per contrast are labelled. ",
  "Axes are free per panel."
)

# Running-enrichment plots: one caption per contrast/kinase, in printing order
gsea_plot_caps <- unlist(lapply(names(gsea_results), function(ct) {
  top_ids <- gsea_results[[ct]]@result |>
    as_tibble() |>
    arrange(pvalue) |>
    head(params$top_genesets) |>
    pull(ID)
  paste0(
    "Running enrichment score of the ", top_ids, " substrate set in contrast ", ct, " (",
    params$analysis_type, "). The x-axis is the rank of the sites ordered by ", inputs$stat_col,
    " from most up- to most down-regulated; the upper panel traces the running enrichment ",
    "score, whose extreme is the enrichment score of the set, the tick marks below mark where ",
    "that kinase's predicted substrates sit in the ranking, and the bottom panel repeats the ",
    "ranking statistic. A curve peaking early with ticks clustered on the left means the ",
    "kinase's substrates are enriched among up-regulated sites."
  )
}), use.names = FALSE)
```

# Results by Contrast {.tabset .tabset-pills}

```{r contrastPlots, results='asis', fig.height=8, eval=has_gsea_results, fig.cap=dotplot_caps}
for (ctr in contrast_ids) {
  cat("\n\n## ", ctr, "\n\n")

  ctr_data <- all_clean |> filter(contrast == ctr)
  n_sig <- sum(ctr_data$p.adjust < 0.1, na.rm = TRUE)
  cat("**Significant kinases (FDR < 0.1):** ", n_sig, "\n\n")

  # Using shared dotplot function
  p <- plot_enrichment_dotplot(
    ctr_data,
    item_col = "kinase",
    fdr_col = "p.adjust",
    title = paste0(params$analysis_type, " - ", ctr),
    subtitle = "Top 30 kinases by FDR"
  )
  print(p)
  cat("\n\n")

  # Significant kinases table
  cat("**Significant Kinases (FDR < 0.1):**\n\n")
  sig_table <- ctr_data |>
    filter(p.adjust < 0.1) |>
    select(kinase, NES, pvalue, FDR = p.adjust, setSize) |>
    arrange(FDR) |>
    mutate(across(where(is.numeric), ~round(.x, 4)))
  print(htmltools::tagList(
    DT::datatable(sig_table,
                  extensions = 'Buttons',
                  options = list(pageLength = 15, scrollX = TRUE,
                                 dom = 'Bfrtip', buttons = c('copy', 'csv', 'excel')),
                  caption = paste0(
                    "Kinases enriched at FDR < 0.1 in contrast ", ctr, ", sorted by FDR. NES ",
                    "is the normalized enrichment score (> 0 predicted substrates enriched ",
                    "among up-regulated sites, < 0 among down-regulated sites), pvalue the ",
                    "GSEA permutation p-value, FDR its Benjamini-Hochberg adjustment and ",
                    "setSize the number of ranked sites assigned to that kinase."
                  ))
  ))
  cat("\n\n")
}
```

# clusterProfiler Dotplots {.tabset .tabset-pills}

### Merged Dotplot

```{r merged-dotplot, fig.width=14, fig.height=10, eval=has_gsea_results, fig.cap=merged_dotplot_cap}
merged <- merge_result(gsea_results)
dotplot(merged, showCategory = 15,
        title = paste(params$analysis_type, "GSEA - Kinase Library (FDR < 0.25)")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Individual Contrasts

```{r individual-dotplots, fig.width=14, fig.height=10, eval=has_gsea_results, fig.cap=individual_dotplot_cap}
plots <- names(gsea_results) |>
  map(function(ct) {
    res <- gsea_results[[ct]]@result
    # Skip if no results
    if (nrow(res) == 0 || sum(!is.na(res$NES)) == 0) {
      return(ggplot() +
        annotate("text", x = 0.5, y = 0.5, label = paste(ct, "\nNo significant kinases")) +
        theme_void() + ggtitle(ct))
    }
    dotplot(gsea_results[[ct]], showCategory = 15, title = ct) +
      theme(axis.text.y = element_text(size = 8))
  })
wrap_plots(plots, ncol = 2)
```

### NES Heatmap

```{r nes-heatmap, fig.width=12, fig.height=10, eval=has_gsea_results, fig.cap=heatmap_cap}
# Using shared heatmap function
plot_enrichment_heatmap(
  all_clean,
  item_col = "ID",
  fdr_col = "p.adjust",
  fdr_filter = 0.25,
  n_top = 30,
  title = paste(params$analysis_type, "- Top Kinases by NES")
)
```

# Volcano Plot

```{r volcano-plot, fig.width=12, fig.height=8, eval=has_gsea_results, fig.cap=volcano_cap}
# Using shared volcano function
plot_enrichment_volcano(
  all_clean,
  item_col = "ID",
  fdr_col = "p.adjust",
  title = paste(params$analysis_type, "- GSEA Volcano Plots")
)
```

# Diagnostics

```{r diagnostic-pvalues, eval=has_gsea_results}
pval_diag <- names(gsea_results) |>
  map_dfr(function(ct) {
    res <- gsea_results[[ct]]@result
    tibble(
      Contrast = ct,
      `Min p-value` = signif(min(res$pvalue, na.rm = TRUE), 3),
      `p < 0.05` = sum(res$pvalue < 0.05, na.rm = TRUE),
      `p < 0.01` = sum(res$pvalue < 0.01, na.rm = TRUE),
      Total = nrow(res)
    )
  })
knitr::kable(
  pval_diag,
  caption = paste0(
    "Distribution of the unadjusted GSEA permutation p-values per contrast (",
    params$analysis_type, "), before FDR adjustment. Total is the number of kinase sets ",
    "tested; a contrast with a real kinase signal shows more sets below the thresholds than ",
    "the roughly 5% and 1% of Total expected by chance."
  )
)
```

# Export All GSEA Plots to PDF

```{r export-gsea-pdf, results='hide', eval=pipeline_mode && has_gsea_results}
# Export all GSEA enrichment plots to PDF using shared function
pdf_file <- file.path(output_dir, paste0("GSEA_plots_", params$analysis_type, ".pdf"))
n_plots <- export_gsea_plots_pdf(gsea_results, pdf_file)
cat("Exported", n_plots, "GSEA plots to:", pdf_file, "\n")
```

```{r export-gsea-pdf-vignette, results='hide', eval=!pipeline_mode && has_gsea_results}
# Vignette mode: skip PDF export
n_plots <- sum(map_int(gsea_results, ~nrow(.x@result)))
message("Vignette mode: PDF export skipped. Would export ", n_plots, " plots.")
```

# GSEA Enrichment Plots {.tabset .tabset-pills}

Showing top `r params$top_genesets` plots per contrast. `r if(pipeline_mode && has_gsea_results) paste0("**All ", n_plots, " plots exported to PDF.**")`

```{r gsea-example-plots, results='asis', fig.width=10, fig.height=6, eval=has_gsea_results, fig.cap=gsea_plot_caps}
for (ct in names(gsea_results)) {
  cat("\n\n### ", ct, " {.tabset}\n\n")

  res <- gsea_results[[ct]]
  top10 <- res@result |>
    as_tibble() |>
    arrange(pvalue) |>
    head(params$top_genesets) |>
    pull(ID)

  for (i in seq_along(top10)) {
    geneset <- top10[i]
    cat("\n\n#### ", geneset, "\n\n")

    row <- res@result |>
      as_tibble() |>
      filter(ID == geneset)
    nes_val <- round(row$NES, 2)
    fdr <- signif(row$p.adjust, 2)

    p <- gseaplot2(res, geneSetID = geneset,
      title = paste0(geneset, " (NES=", nes_val, ", FDR=", fdr, ")"))
    print(p)
    cat("\n\n")
  }
}
```

# All Results

```{r allResults, eval=has_gsea_results}
# All kinases across all contrasts
all_results_dt <- names(gsea_results) |>
  map_dfr(function(ct) {
    gsea_results[[ct]]@result |>
      as_tibble() |>
      mutate(contrast = ct) |>
      select(contrast, kinase = ID, NES, pvalue, FDR = p.adjust, setSize)
  }) |>
  arrange(contrast, FDR) |>
  mutate(across(where(is.numeric), ~round(.x, 4)))

DT::datatable(all_results_dt,
  filter = "top",
  extensions = 'Buttons',
  options = list(pageLength = 15, scrollX = TRUE,
                 dom = 'Bfrtip', buttons = c('copy', 'csv', 'excel')),
  caption = paste0(
    "Complete GSEA result table: every tested kinase substrate set in every contrast (",
    params$analysis_type, "), sorted by contrast and FDR, including the non-significant ones. ",
    "NES is the normalized enrichment score (> 0 predicted substrates enriched among ",
    "up-regulated sites, < 0 among down-regulated sites), pvalue the permutation p-value, FDR ",
    "its Benjamini-Hochberg adjustment and setSize the number of ranked sites assigned to the ",
    "kinase. Use the column filters to search a kinase."
  ))
```

# Export Results

The result workbook and the RDS this report renders from were written by the
computation step, `prophosqua::compute_kinaselib_gsea()`, before this render.
Both are written whether or not anything was enriched, so the pipeline can
declare them as outputs and an empty enrichment is a recorded fact rather than a
missing file.

```{r exportedFiles}
stem <- file.path(output_dir, paste0("KinaseLib_GSEA_", params$analysis_type))
exported <- paste0(stem, c(".xlsx", ".rds"))

knitr::kable(
  tibble(
    Output = c("Excel results", "RDS object"),
    File = basename(exported),
    Size = ifelse(
      file.exists(exported),
      paste(round(file.size(exported) / 1024, 1), "KB"),
      "not found"
    )
  ),
  caption = paste0(
    "Files the computation step wrote next to this report: the Excel workbook with all ",
    "kinase results, a significant-only sheet and the per-contrast summary, and the RDS with ",
    "the full GSEA result objects and the tables shown above."
  )
)
```

## Session Info

```{r session-info}
sessionInfo()
```
